---
title: "Clustering clean merged and non-merged reads separately"
author: "J. Ignacio Lucas Lledó"
date: "18/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In folder `2022-01-10` I failed to produce the clustering of reads,
because the excess and type of data requires too much RAM. Here I
will repeat the filtering to correct the fact that an absolute limit
of expected errors was invariably imposed on reads of differing lengths.
Then, I will keep merged and non-merged reads separate after de-replication,
hoping that their separate clustering will be managable and still
meaningful. The assumption is that non-merged reads with enough quality
must correspond to different loci, rather than having failed to merge.

## Filtering by an error rate of 1 in 100

```{bash FilteringAgain}
MERGED=../2022-01-10/merged
PAIRED=../2022-01-10/paired
if [ ! -d merged.clean ]; then mkdir merged.clean; fi
if [ ! -d paired.clean ]; then mkdir paired.clean; fi
if [ ! -d logs ]; then mkdir logs; fi
if [ ! -e filtering.stats.txt ]; then
   # I use my parallelization, because --fastq_filter is not parallelized.
   NUMSAMPLES=$(ls -1 $MERGED | wc -l)
   NUMTHREADS=40
   while [ $(( $NUMSAMPLES % $NUMTHREADS )) -ne 0 ]; do
      NUMTHREADS=$(( NUMTHREADS + 1))
   done
   NUMBATCHES=$(( NUMSAMPLES / NUMTHREADS))
   for i in $(seq 1 $NUMBATCHES); do
      LAST=$((  NUMTHREADS * i ))
      for sample in $(ls -1 $MERGED | cut -d "." -f 1 | head -n $LAST | tail -n $NUMTHREADS); do
         if [ ! -e merged.clean/$sample.fq.gz ]; then
            vsearch --fastq_filter $MERGED/$sample.fq.gz \
                    --fastqout - \
                    --log logs/$sample.merged.filter.log \
                    --fastq_minlen 36 \
                    --fastq_maxee_rate 0.01 | \
            gzip > merged.clean/$sample.fq.gz &
         fi
      done
      wait
   done
   for i in $(seq 1 $NUMBATCHES); do
      LAST=$((  NUMTHREADS * i ))
      for sample in $(ls -1 $MERGED | cut -d "." -f 1 | head -n $LAST | tail -n $NUMTHREADS); do
         if [ ! -e paired.clean/$sample.R1.fq.gz ]; then
            vsearch --fastq_filter $PAIRED/$sample.R1.fq.gz \
                    --fastqout - \
                    --log logs/$sample.paired.filter.log \
                    --fastq_maxee_rate 0.01 | \
            gzip > paired.clean/$sample.R1.fq.gz &
         fi
      done
      wait
   done

   gawk 'BEGIN{
      print "Sample\tMerged.Kept\tMerged.Lost\tPaired.Kept\tPaired.Lost"
   }(/sequences kept/){
      split(FILENAME, A, /[./]/)
      SAMPLE = A[2]
      SAMPLELIST[SAMPLE] = 1
      TYPE = A[3]
      KEPT[SAMPLE, TYPE] = $1
      LOST[SAMPLE, TYPE] = $8
   }END{
      for (sample in SAMPLELIST) {
         print sample "\t" KEPT[sample, "merged"] "\t" \
         LOST[sample, "merged"] "\t" KEPT[sample, "paired"] "\t" \
         LOST[sample, "paired"]
      }
   }' logs/*.filter.log > filtering.stats.txt
fi
```

```{r filteringStats}

library(tidyr)
library(gridExtra)
library(ggplot2)
library(RColorBrewer)

LengthHist <- read.table('../2022-01-10/merged.stats.txt', header = TRUE)
LongHist <- pivot_longer(LengthHist, 2:281, names_to = 'Sample', values_to = 'NumReads')
Filters <- lapply(unique(substr(LongHist$Sample, 1, 3)),
                  function(x) startsWith(LongHist$Sample, x))
names(Filters) <- unique(substr(LongHist$Sample, 1, 3))

Merging <- data.frame(
   Sample = names(LengthHist)[2:281],
   Merged = colSums(LengthHist[1:(dim(LengthHist)[1] - 1), 2:281]),
   Paired = as.numeric(LengthHist[dim(LengthHist)[1], 2:281]),
   Total  = colSums(LengthHist[, 2:281])
)

Filtering <- read.table('filtering.stats.txt', header = TRUE, as.is = TRUE)
f_175 <- startsWith(Filtering$Sample, '175')
Filtering[f_175, 'Sample'] <- paste0('X', Filtering[f_175, 'Sample'])
row.names(Filtering) <- Filtering$Sample
Filtering <- Filtering[row.names(Merging),]
Palette <- brewer.pal(4, 'Paired')
barplot(t(as.matrix(Filtering[order(Merging$Total, decreasing = TRUE), 2:5])),
        xaxt = 'n', border = NA, space = 0,
        col = Palette)
legend(180, 2e+07, c('Merged kept', 'Merged lost', 'Paired kept', 'Paired lost'),
       fill = Palette)
```

## De-replication

```{bash dereplication}
MERGED=../2022-01-10/merged
PAIRED=../2022-01-10/paired
if [ ! -d merged.derep ]; then mkdir merged.derep; fi
if [ ! -d paired.derep ]; then mkdir paired.derep; fi
if [ ! -e paired.derep.stats.txt ]; then
   NUMSAMPLES=$(ls -1 $MERGED | wc -l)
   NUMTHREADS=40
   while [ $(( $NUMSAMPLES % $NUMTHREADS )) -ne 0 ]; do
      NUMTHREADS=$(( NUMTHREADS + 1))
   done
   NUMBATCHES=$(( NUMSAMPLES / NUMTHREADS))
   for i in $(seq 1 $NUMBATCHES); do
      LAST=$((  NUMTHREADS * i ))
      for sample in $(ls -1 $MERGED | cut -d "." -f 1 | head -n $LAST | tail -n $NUMTHREADS); do
         if [ ! -e merged.derep/$sample.fa ]; then
            vsearch --derep_fulllength merged.clean/$sample.fq.gz \
                    --gzip_decompress \
                    --sizeout \
                    --fasta_width 0 \
                    --output - |
            gawk -v SAMPLE=$sample '(/^>/){
               print $1 ";sample=" SAMPLE
            }(/^[^>]/){
               print $1
            }' > merged.derep/$sample.fa 2> logs/$sample.merged.derep.log &
         fi
      done
      wait
   done
   for i in $(seq 1 $NUMBATCHES); do
      LAST=$((  NUMTHREADS * i ))
      for sample in $(ls -1 $MERGED | cut -d "." -f 1 | head -n $LAST | tail -n $NUMTHREADS); do
         if [ ! -e paired.derep/$sample.fa ]; then
            vsearch --derep_fulllength paired.clean/$sample.R1.fq.gz \
                    --gzip_decompress \
                    --sizeout \
                    --fasta_width 0 \
                    --output - |
            gawk -v SAMPLE=$sample '(/^>/){
               print $1 ";sample=" SAMPLE
            }(/^[^>]/){
               print $1
            }' > paired.derep/$sample.fa 2> logs/$sample.paired.derep.log &
         fi
      done
      wait
   done
   # The following reads the de-replicated fasta files and writes a
   # matrix with the histograms of multiplicity (number of sequences
   # observed x number of times) for all samples.
   gawk 'BEGIN{
      MAXREP=30000
      SAMPLEINDEX=1
   }(FNR == 1){
      split(FILENAME, PATH, /[/.]/)
      SAMPLE = PATH[3]
      SAMPLELIST[SAMPLEINDEX] = SAMPLE
      SAMPLEINDEX++
      HEADER = HEADER "\t" SAMPLE
   }(/^>/){
      split($1, A, /;/)
      split(A[length(A) - 1], B, /=/)
      F[SAMPLE, B[2]]++
      if (B[2] > MAXREP) MAXREP = B[2]
   }END{
      print HEADER
      for (i = 1; i <= MAXREP; i++) {
         LINE = i
         for (j = 1; j <= length(SAMPLELIST); j++) {
            LINE = LINE "\t" F[SAMPLELIST[j], i] + 0 
         }
         print LINE
      }
   }' merged.derep/*.fa > merged.derep.stats.txt
   
   gawk 'BEGIN{
      MAXREP=30000
      SAMPLEINDEX=1
   }(FNR == 1){
      split(FILENAME, PATH, /[/.]/)
      SAMPLE = PATH[3]
      SAMPLELIST[SAMPLEINDEX] = SAMPLE
      SAMPLEINDEX++
      HEADER = HEADER "\t" SAMPLE
   }(/^>/){
      split($1, A, /;/)
      split(A[length(A) - 1], B, /=/)
      F[SAMPLE, B[2]]++
      if (B[2] > MAXREP) MAXREP = B[2]
   }END{
      print HEADER
      for (i = 1; i <= MAXREP; i++) {
         LINE = i
         for (j = 1; j <= length(SAMPLELIST); j++) {
            LINE = LINE "\t" F[SAMPLELIST[j], i] + 0 
         }
         print LINE
      }
   }' paired.derep/*.fa > paired.derep.stats.txt
fi
```

```{r plotDerepStats, fig.width=10, fig.height=10}
Derep.m <- read.table('merged.derep.stats.txt', header = TRUE)
# Pool all counts above a multiplicity of 6000 in the same bin:
Derep.m[6001, ] <- lapply(Derep.m, function(x, END) sum(x[6001:END]), END = dim(Derep.m)[1])
Derep.m <- Derep.m[1:6001,] + 1
Derep.m$Multiplicity <- 1:6001
Derep.m.long <- pivot_longer(Derep.m, 1:280, names_to = 'Sample', values_to = 'NumReads')
Filters <- lapply(unique(substr(Derep.m.long$Sample, 1, 3)),
                  function(x) grepl(x, Derep.m.long$Sample))
names(Filters) <- unique(substr(Derep.m.long$Sample, 1, 3))
plots <- lapply(names(Filters),
                function(lake) {
                   ggplot(Derep.m.long[Filters[[lake]], ],
                          aes(x = Multiplicity, y = NumReads, color = Sample)) +
                      geom_line() + guides(color = 'none') + ggtitle(lake) +
                      xlab('Multiplicity') + ylab('Number of reads') +
                      scale_y_log10() + scale_x_log10()
                })
grid.arrange(grobs = plots, nrow = 3)

Derep.p <- read.table('paired.derep.stats.txt', header = TRUE)
# Pool all counts above a multiplicity of 6000 in the same bin:
Derep.p[6001, ] <- lapply(Derep.p, function(x, END) sum(x[6001:END]), END = dim(Derep.p)[1])
Derep.p <- Derep.p[1:6001,] + 1
Derep.p$Multiplicity <- 1:6001
Derep.p.long <- pivot_longer(Derep.p, 1:280, names_to = 'Sample', values_to = 'NumReads')
Filters <- lapply(unique(substr(Derep.p.long$Sample, 1, 3)),
                  function(x) grepl(x, Derep.p.long$Sample))
names(Filters) <- unique(substr(Derep.p.long$Sample, 1, 3))
plots <- lapply(names(Filters),
                function(lake) {
                   ggplot(Derep.p.long[Filters[[lake]], ],
                          aes(x = Multiplicity, y = NumReads, color = Sample)) +
                      geom_line() + guides(color = 'none') + ggtitle(lake) +
                      xlab('Multiplicity') + ylab('Number of reads') +
                      scale_y_log10() + scale_x_log10()
                })
grid.arrange(grobs = plots, nrow = 3)
```

## Clustering

```{bash clustering}
if [ ! -e merged.centroids.fa ]; then
   if [ ! -e merged.sorted.fa ]; then
      cat merged.derep/*.fa | \
      vsearch --sortbylength - \
              --minseqlength 100 \
              --fasta_width 0 \
              --output merged.sorted.fa
   fi
   vsearch --cluster_smallmem merged.sorted.fa \
           --centroids merged.centroids.fa \
           --id 0.97 \
           --sizein \
           --sizeout \
           --strand plus \
           --threads 1 \
           --log logs/merged.clustering.log \
           --fasta_width 0
fi
if [ ! -e paired.centroids.fa ]; then
   if [ ! -e paired.sorted.fa ]; then
      cat paired.derep/*.fa | \
      vsearch --sortbylength - \
              --minseqlength 100 \
              --fasta_width 0 \
              --output paired.sorted.fa
   fi
   vsearch --cluster_smallmem paired.sorted.fa \
           --centroids paired.centroids.fa \
           --id 0.97 \
           --sizein \
           --sizeout \
           --strand plus \
           --threads 1 \
           --log logs/paired.clustering.log \
           --fasta_width 0
fi
#           --consout consensus.fa \
#           --otutabout clustering.txt \
#           --clusterout_id \
```

## Session Information

```{r sessionInfo}
sessionInfo()
```
