---
title: "Index hopping"
author: "J. Ignacio Lucas Lledó"
date: "13/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We have now two batches of reads, because the sequencing center decided
to run the machine a second time. Here I add the number of reads in the
second batch to the table of sample features and I try to model the amount
of index hopping among the index combinations where we can observe the
misassignment of reads.

```{bash numReads}
if [ ! -e TotalNumReads.txt ]; then
   FASTQDIR1='../../data/fastq'
   FASTQDIR2='../../data/fastr'
   if [ ! -e SampleIndexSeq.tsv ]; then cp ../2021-10-13/SampleIndexSeq.tsv ./; fi
   
   SAMPLE=( $(gawk '(NR > 1){gsub(/\"/,""); print $1}' SampleIndexSeq.tsv | sort) )
   THREADS=24
   for i in $(seq 0 $THREADS $(( ${#SAMPLE[@]} - 1 )) ); do
      for j in $(seq $i $(($i + $THREADS - 1)) ); do
         echo -e ${SAMPLE[$j]} "\t" \
            $(echo $(gunzip -c $FASTQDIR1/${SAMPLE[$j]}*_R1_001.fastq.gz 2> /dev/null | wc -l) "/4" | bc) "\t" \
            $(echo $(gunzip -c $FASTQDIR2/${SAMPLE[$j]}*_R1_001.fastq.gz 2> /dev/null | wc -l) "/4" | bc) \
            > $(printf "z_%s.txt" ${SAMPLE[$j]}) &
      done
      wait
   done
   for k in $(seq $(( $j + 1)) $(( ${#SAMPLE[@]} - 1)) ); do
      echo -e ${SAMPLE[$k]} "\t" \
         $(echo $(gunzip -c $FASTQDIR1/${SAMPLE[$k]}*_R1_001.fastq.gz 2> /dev/null | wc -l) "/4" | bc) "\t" \
         $(echo $(gunzip -c $FASTQDIR2/${SAMPLE[$k]}*_R1_001.fastq.gz 2> /dev/null | wc -l) "/4" | bc) \
         > $(printf "z_%s.txt" ${SAMPLE[$k]}) &
   done
   wait
   cat z_*.txt | sort -k 1,1 > TotalNumReads.txt
   rm z*
fi

if [ ! -e Features.RData ]; then
   cp ../2021-11-23/Features.RData ./
fi
```

```{r data}
library('ggplot2')
load('Features.RData')
TotalNumReads <- read.table('TotalNumReads.txt',
                            col.names = c('Sample','numReads1', 'numReads2'),
                            row.names = 1)
Features$numReads1 <- TotalNumReads[row.names(Features), 'numReads1']
Features$numReads2 <- TotalNumReads[row.names(Features), 'numReads2']
Features$meanNumReads <- (Features$numReads1 + Features$numReads2) / 2
# Coefficient of variation between the two runs: CV(x)=sd(x)/mean(x).
# A downwardly biased estimation of sd of two numbers: abs(x - y)/sqrt(2)
Features$CVnumReads <- abs(Features$numReads1 - Features$numReads2) / (sqrt(2) * Features$meanNumReads)
stopifnot(all.equal(Features$numReads, Features$numReads1))
Features <- Features[, -(which(names(Features) == 'numReads'))]

f <- Features$i7 != 'N729'
ggplot(Features[f,], aes(x = meanNumReads, y = CVnumReads, color = isReal)) +
  geom_point() + scale_y_log10() + ylab('CV number of reads between runs') +
  xlab('Mean number of reads')

ggplot(data = Features, mapping = aes(x = numReads1, y = numReads2, color = isReal)) +
  geom_point() + scale_y_log10() + scale_x_log10() +
  xlab('First sequencing run') + ylab('Second sequencing run') + ggtitle('Number of reads per sample')
```

The sequencing run does not affect much the number of reads. Even the amount of
index hopping is the same between technical replicates of the sequencing.

## Determinants of index hopping

```{r heatmap}
ggplot(data = Features,
       mapping = aes(x = i7, y = i5, fill = meanNumReads)) +
   geom_tile() + geom_hline(yintercept = 8.5) + geom_vline(xintercept = 12.5) +
   scale_fill_gradient(trans = 'log',
                       breaks = c(1,10,100,1000,10000,100000,1000000,10000000)) +
   theme(axis.text.x = element_text(angle = 90)) +
   scale_y_discrete(limits = rev) +
   annotate(geom = 'text', x = c(6,18,6,18), y=c(12,12,4,4), label=c('A','B','C','D'))
```

There are different kinds of samples, in terms of their expected amount of index
hopping:

1. 278 real samples identified with i5 and i7 indices that are both shared
with other real samples.
2. Two real samples identified with i5 and i7 indices not shared with any
other sample.
3. And mock samples all the reads of which come from index hopping.

Among the mock samples, those with i7 index N729 stick out for having extremely
low numbers of reads: zero or one. This is the only index that was not used
at all. Because its tube was never opened, molecules with this index must
have been absent in the pool. The few reads reporting N729 can only come from
technical artifacts, such as errors in reading the index sequence. Thus, most
of the index hopping must be due not to sequencing errors, but to real molecules.

The presence of free adapters is supposed to be the main cause of index
hopping. After having cleaned up the pools and after size selection, the
amount of free adapters in the pool must have been negligible. Fragmentation
of the DNA fragments in the libraries can also produce free adapters, and
this is the reason why they recommend libraries to be stored at -20 C.

In any case, the number of reads missassigned to an index must be proportional
to the number of fragments in the pool actually labeled with that index. This
is quite apparent in the figure above, but can be made more explicit. Furthermore
I wonder if missassignment to an erroneous i5 is just as probable as missassignment
to an erroneous i7. 

```{r hoppingNeighbours}
# Number of real samples sharing one or the other index (excluding itself, if real):
Features$i5neiNum <- sapply(1:384, function(x){
  sum(Features$isReal & Features$i5 == Features[x, 'i5'])
}) - Features$isReal
Features$i7neiNum <- sapply(1:384, function(x){
  sum(Features$isReal & Features$i7 == Features[x, 'i7'])
}) - Features$isReal
Features$neiNum     <- Features$i5neiNum + Features$i7neiNum
Features$neiNumProd <- Features$i5neiNum * Features$i7neiNum

f <- ! Features$isReal & Features$i7 != 'N729'
summary(lm(meanNumReads ~ neiNum, data = Features[f, ]))
summary(lm(meanNumReads ~ i5neiNum + i7neiNum, data = Features[f,]))
summary(lm(meanNumReads ~ i5neiNum * i7neiNum, data = Features[f,]))
summary(lm(meanNumReads ~ neiNumProd, data = Features[f,]))

# Number of real samples sharing one index or the other and in the same plate.
Features$i5neiNumPlate <- sapply(1:384, function(x){
  sum(Features$isReal &
        Features$i5 == Features[x, 'i5'] &
        Features$plate == Features[x, 'plate'])
}) - Features$isReal
Features$i7neiNumPlate <- sapply(1:384, function(x){
  sum(Features$isReal &
        Features$i7 == Features[x, 'i7'] &
        Features$plate == Features[x, 'plate'])
}) - Features$isReal
Features$neiNumPlate     <- Features$i5neiNumPlate + Features$i7neiNumPlate
Features$neiNumPlateProd <- Features$i5neiNumPlate * Features$i7neiNumPlate

f <- ! Features$isReal & Features$i7 != 'N729'
summary(lm(meanNumReads ~ neiNumPlate, data = Features[f,]))
summary(lm(meanNumReads ~ i5neiNumPlate + i7neiNumPlate, data = Features[f,]))
summary(lm(meanNumReads ~ i5neiNumPlate * i7neiNumPlate, data = Features[f,]))
summary(lm(meanNumReads ~ neiNumPlateProd, data = Features[f,]))

ggplot(data = Features[f,], mapping = aes(x = neiNumProd, y = meanNumReads, color = plate)) +
  geom_point() + xlab('Product of number of real neighbours') + ylab('Number of missassigned reads')
```

It is not the number of real samples *in the same plate*, but the product
or interaction between the number of real samples sharing one index and
the number sharing the other *in the whole experiment* what determines
better the amount of index hopping.

```{r hoppingReads}
# Number of reads with same index i5 or i7
Features$i5neiReads <- sapply(1:384, function(x) {
  sum(Features[Features$isReal & Features$i5 == Features[x,'i5'], 'meanNumReads'])
}) - Features$isReal * Features$meanNumReads
Features$i7neiReads <- sapply(1:384, function(x) {
  sum(Features[Features$isReal & Features$i7 == Features[x,'i7'], 'meanNumReads'])
}) - Features$isReal * Features$meanNumReads
Features$neiReads     <- Features$i5neiReads + Features$i7neiReads
Features$neiReadsProd <- Features$i5neiReads * Features$i7neiReads

f <- ! Features$isReal & Features$i7 != 'N729'
summary(lm(meanNumReads ~ neiReads, data = Features[f,]))
summary(lm(meanNumReads ~ i5neiReads + i7neiReads, data = Features[f,]))
summary(lm(meanNumReads ~ i5neiReads * i7neiReads, data = Features[f,]))
summary(lm(meanNumReads ~ neiReadsProd, data = Features[f,]))

ggplot(data = Features[f,], mapping = aes(x = neiReadsProd, y = meanNumReads, color = i7neiNum)) +
  geom_point() + geom_smooth(method = 'lm') + scale_y_log10() + scale_x_log10() +
  xlab('Product of numbers of index neighbours') + ylab('Number of missassigned reads') +
  ggtitle('Index hopping')
```

## Conclusions

I have modelled the number of reads in non real samples. That is, the number
of misassigned reads where we can count them. I expected misassignments to be
due to independent events involving a misreading of either index i5 or index i7.
However, the sum of *i5-neighbours* and *i7-neighbours* is not as predictive
of the amount of misassignment as their product. This points at an interaction
between the number of samples (or reads) sharing the i5 and the number sharing
the i7 indices. If I understand well, this means that errors do not only come
from misreading one of the two indices, but often from misreading both.

It is also notable that the cross-contamination is not limited to the plate,
but affects the whole pool together. This suggests that the problem is **not**
an erroneous manipulation of the indices during library preparation, but something
that happens after pooling all samples together.

## Prediction of misassigned reads in real samples

```{r prediction}
f <- ! Features$isReal & Features$i7 != 'N729'
M0 <- lm(meanNumReads ~ i5neiReads * i7neiReads, data = Features[f,])
Predicted <- predict(M0, Features[Features$isReal, c('i5neiReads', 'i7neiReads')], se.fit = TRUE)
plot(density(Predicted$fit / Features[names(Predicted$fit), 'meanNumReads']),
     xlab = 'Expected fraction of index hopping', main = '')
```

The only sample where we expect a proportion of wrongly assigned reads
is the one with fewest reads: SUO095, which looked ok in the electrophoresis
but turned out with only 167111 reads between the two sequencing runs.

```{r sessionInfo}
sessionInfo()
```

