---
title: "Quality control of fastq files"
author: "Mar Llaberia Robledillo & J. Ignacio Lucas Lledó"
date: "10/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are two sequencing runs, each with the two fastq files
from each of the 280 true samples and the 94 mock ones. In all
they are 1496 fastq files. To visualize quality parameters of
all them is too much.

```{r script}
library(fastqcr)
run1 <- '../../data/fastq'
run2 <- '../../data/fastr'

if (! file.exists('report1')) {
  fastqc(fq.dir = run1, qc.dir = "reports1", 
         threads = 40, fastqc.path = "/usr/local/bin/fastqc")
}

if (! file.exists('report2')) {
  fastqc(fq.dir = run2, qc.dir = "reports2",
         threads = 40, fastqc.path = "/usr/local/bin/fastqc")
}

qc.files1 <- list.files(path = "reports1", all.files = FALSE,
                        full.names = TRUE, pattern = "zip")
# Sample names can be extracted from the file names
samples1 <- sapply(strsplit(qc.files1, "[/_]"), '[[', 2)

qc.files2 <- list.files(path = 'reports2', all.files = FALSE,
                        full.names = TRUE, pattern = 'zip')
# Some fake samples (HOP) got so little spurious reads that appear in
# only one of the two sequencing runs. I will limit to the subset of
# samples with fastq files from both sequencing runs.
samples2 <- sapply(strsplit(qc.files2, "[/_]"), '[[', 2)
samples <- intersect(samples1, samples2)
rm(samples1, samples2)

# The lists of fastq files from the two sequencing runs are not
# "parallel". I will use the common samples to filter the lists of files:
qc.files1 <- grep(paste0('(', paste(samples, collapse = '|'), ')'), qc.files1, value = TRUE)
qc.files2 <- grep(paste0('(', paste(samples, collapse = '|'), ')'), qc.files2, value = TRUE)
stopifnot(all.equal(substring(qc.files1, 10), substring(qc.files2, 10)))
```

To read collections of reports is slow, and we don't want to include
all samples in a collection, because then the plots would not be
visible. One way to split samples in meaningful groups would be to
use the 3-letters prefix of their names, which indicate the lake
of origin, I think.

```{r collections}
table(substr(samples, 1, 3))
# filters for individual lakes:
f_lake <- lapply(unique(substr(samples, 1, 3)),
                 function(x) grepl(x, qc.files1))
# just to check
stopifnot(all.equal(as.numeric(table(substr(samples, 1, 3))),
                    sapply(f_lake, sum)/2))
names(f_lake) <- unique(substr(samples, 1, 3))

# And filters for forward and reverse reads:
f_R1  <- grepl('_R1_', qc.files1)
f_R2  <- grepl('_R2_', qc.files1)

# I create 4 lists of collections for each lake:
coll_R1.run1 <- lapply(unique(substr(samples, 1, 3)),
                       function(x) {
                         suppressMessages(qc_read_collection(
                           files = qc.files1[f_lake[[x]] & f_R1],
                           sample_names = grep(x, samples, value = TRUE),
                           modules = 'all',
                           verbose = FALSE
                         ))
                       })
names(coll_R1.run1) <- unique(substr(samples, 1, 3))

coll_R2.run1 <- lapply(unique(substr(samples, 1, 3)),
                       function(x) {
                         suppressMessages(qc_read_collection(
                           files = qc.files1[f_lake[[x]] & f_R2],
                           sample_names = grep(x, samples, value = TRUE),
                           modules = 'all',
                           verbose = FALSE
                         ))
                       })
names(coll_R2.run1) <- unique(substr(samples, 1, 3))

coll_R1.run2 <- lapply(unique(substr(samples, 1, 3)),
                       function(x) {
                         suppressMessages(qc_read_collection(
                           files = qc.files2[f_lake[[x]] & f_R1],
                           sample_names = grep(x, samples, value = TRUE),
                           modules = 'all',
                           verbose = FALSE
                         ))
                       })
names(coll_R1.run2) <- unique(substr(samples, 1, 3))

coll_R2.run2 <- lapply(unique(substr(samples, 1, 3)),
                       function(x) {
                         suppressMessages(qc_read_collection(
                           files = qc.files2[f_lake[[x]] & f_R2],
                           sample_names = grep(x, samples, value = TRUE),
                           modules = 'all',
                           verbose = FALSE
                         ))
                       })
names(coll_R2.run2) <- unique(substr(samples, 1, 3))
```

In order to reduce the size of the report and to be able to compare
forward and reverse reads as well as the two sequencing runs, I will
group plots by lake. Not all modules produce meaningful plots when the
collection includes several reports. For example, the module "Per base
sequence content" would try to produce as many plots as reports, instead
of summarizing them in a single plot.

## Per base sequence quality

```{r perBaseQ, fig.width=10, fig.height=10}
library('gridExtra')
library('ggplot2')
for (lake in unique(substr(samples, 1, 3))) {
  whatPlot <- 'Per base sequence quality'
  p1 <- qc_plot_collection(coll_R1.run1[[lake]], modules = whatPlot)
  p2 <- qc_plot_collection(coll_R2.run1[[lake]], modules = whatPlot)
  p3 <- qc_plot_collection(coll_R1.run2[[lake]], modules = whatPlot)
  p4 <- qc_plot_collection(coll_R2.run2[[lake]], modules = whatPlot)
  p1 <- p1 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 1')
  p2 <- p2 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 1')
  p3 <- p3 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 2')
  p4 <- p4 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 2')
  if (sum(f_lake[[lake]]) > 44) {
    p1 <- p1 + guides(color = 'none')
    p2 <- p2 + guides(color = 'none')
    p3 <- p3 + guides(color = 'none')
    p4 <- p4 + guides(color = 'none')
  }
  grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
}
```

## Per sequence quality scores

```{r perSeqQ, fig.width=10, fig.height=10}
for (lake in unique(substr(samples, 1, 3))) {
  whatPlot <- 'Per sequence quality scores'
  p1 <- qc_plot_collection(coll_R1.run1[[lake]], modules = whatPlot)
  p2 <- qc_plot_collection(coll_R2.run1[[lake]], modules = whatPlot)
  p3 <- qc_plot_collection(coll_R1.run2[[lake]], modules = whatPlot)
  p4 <- qc_plot_collection(coll_R2.run2[[lake]], modules = whatPlot)
  p1 <- p1 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 1')
  p2 <- p2 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 1')
  p3 <- p3 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 2')
  p4 <- p4 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 2')
  if (sum(f_lake[[lake]]) > 44) {
    p1 <- p1 + guides(color = 'none')
    p2 <- p2 + guides(color = 'none')
    p3 <- p3 + guides(color = 'none')
    p4 <- p4 + guides(color = 'none')
  }
  grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
}
```

## Per sequence GC content

```{r perSeqGC, fig.width=10, fig.height=10}
for (lake in unique(substr(samples, 1, 3))) {
  whatPlot <- 'Per sequence GC content'
  p1 <- qc_plot_collection(coll_R1.run1[[lake]], modules = whatPlot)
  p2 <- qc_plot_collection(coll_R2.run1[[lake]], modules = whatPlot)
  p3 <- qc_plot_collection(coll_R1.run2[[lake]], modules = whatPlot)
  p4 <- qc_plot_collection(coll_R2.run2[[lake]], modules = whatPlot)
  p1 <- p1 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 1')
  p2 <- p2 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 1')
  p3 <- p3 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 2')
  p4 <- p4 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 2')
  if (sum(f_lake[[lake]]) > 44) {
    p1 <- p1 + guides(color = 'none')
    p2 <- p2 + guides(color = 'none')
    p3 <- p3 + guides(color = 'none')
    p4 <- p4 + guides(color = 'none')
  }
  grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
}
```

## Per base N content

```{r perBaseN, fig.width=10, fig.height=10}
for (lake in unique(substr(samples, 1, 3))) {
  whatPlot <- 'Per base N content'
  p1 <- qc_plot_collection(coll_R1.run1[[lake]], modules = whatPlot)
  p2 <- qc_plot_collection(coll_R2.run1[[lake]], modules = whatPlot)
  p3 <- qc_plot_collection(coll_R1.run2[[lake]], modules = whatPlot)
  p4 <- qc_plot_collection(coll_R2.run2[[lake]], modules = whatPlot)
  p1 <- p1 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 1')
  p2 <- p2 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 1')
  p3 <- p3 + ggtitle(lake, subtitle = 'Forward reads, sequencing run 2')
  p4 <- p4 + ggtitle(lake, subtitle = 'Reverse reads, sequencing run 2')
  if (sum(f_lake[[lake]]) > 44) {
    p1 <- p1 + guides(color = 'none')
    p2 <- p2 + guides(color = 'none')
    p3 <- p3 + guides(color = 'none')
    p4 <- p4 + guides(color = 'none')
  }
  grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
}
```

## Conclusions
We have a lot of data. The quality is good. I don't know why the sequencing
center repeated the sequencing run. We can be conservative when filtering.

## Session Information

```{r sessionInfo}
sessionInfo()
```