---
title: "Index sequences"
author: "J. Ignacio Lucas Lledó"
date: "13/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is only intended to produce in a clear, reproducible way the assignation
of Nextera indices (v2) to samples, according to the positions of samples in the four
96-well plates. Sample positions are taken from a spreadsheet created by Alan (`Disseny Plaques.xlsx`).

The index sequences are taken from an Illumina Document # 1000000002694 v16 (April 2021, Illumina),
called  "Illumina Adapter Sequences" (pp 29-30). The *i7* indices have two sequences: the real onw in the adapter
and the reverse complementary, which is the one actually read by the machine. Thus, the second column
in *i7* is the sequence to be specified in a sample sheet. Regarding *i5* there are three columns:
the real sequence twice and the reverse complement, because some machines read the forward and others the
reverse strand of the adapter with the *i5* index. In the case of NextSeq2000, as for NovaSeq6000, it
is the sequence as it appears in the adapter what is read. However, the instructions to fill up the
sequencing request form say: "Double-check the sequence is NOT the reverse complement of the index".
Thus, I will report only the sequences as they appear in the adapters.

```{r data}
A  <- read.table('../../data/plate_A.tsv', header = TRUE, as.is = TRUE)
B  <- read.table('../../data/plate_B.tsv', header = TRUE, as.is = TRUE)
C  <- read.table('../../data/plate_C.tsv', header = TRUE, as.is = TRUE)
D  <- read.table('../../data/plate_D.tsv', header = TRUE, as.is = TRUE)
i7 <- read.table('../../data/i7_indices.tsv', col.names = c('id', 'inAdapter', 'inSheet'), row.names = 1, as.is = TRUE)
i5 <- read.table('../../data/i5_indices.tsv', col.names = c('id', 'inAdapter', 'inSheet1', 'inSheet2'), row.names = 1, as.is = TRUE)
stopifnot(all.equal(rownames(A), rownames(B)))
stopifnot(all.equal(rownames(C), rownames(D)))
stopifnot(all.equal(colnames(A), colnames(C)))
stopifnot(all.equal(colnames(B), colnames(D)))

AllSamples <- as.matrix(rbind(cbind(A,B), cbind(C,D)))

IndexSample <- do.call(
  rbind,
  lapply(colnames(AllSamples), function(x) {
    t(sapply(rownames(AllSamples), function(y, i=x) c(AllSamples[y, i], i, i7[i, 'inAdapter'], y, i5[y, 'inAdapter']), USE.NAMES = FALSE))
  })
)

colnames(IndexSample) <- c('Sample','i7','i7seq','i5','i5seq')
IndexSample <- IndexSample[! is.na(IndexSample[,'Sample']),]
write.table(IndexSample, file = 'SampleIndexSeq.tsv', sep = '\t', row.names = FALSE)
```

## Sesion Info

```{r sessionInfo}
sessionInfo()
```

