---
title: "Index sequences"
author: "J. Ignacio Lucas Lledó"
date: "13/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

This document is only intended to produce in a clear, reproducible way the assignation
of Nextera indices (v2) to samples, according to the positions of samples in the four
96-well plates. Sample positions are taken from a spreadsheet created by Alan (`Disseny Plaques.xlsx`).

The index sequences are taken from an Illumina Document # 1000000002694 v16 (April 2021, Illumina),
called  "Illumina Adapter Sequences" (pp 29-30). The *i7* indices have two sequences: the real one in the adapter
and the reverse complementary, which is the one actually read by the machine. Thus, the second column
in *i7* is the sequence to be specified in a sample sheet. Regarding *i5* there are three columns:
the real sequence twice and the reverse complement, because some machines read the forward and others the
reverse strand of the adapter with the *i5* index. In the case of NextSeq2000, as for NovaSeq6000, it
is the sequence as it appears in the adapter what is read. Indeed, the instructions to fill up the
sequencing request form say: "Double-check the sequence is NOT the reverse complement of the index".
Thus, I will report only the sequences as they appear in the adapters.

```{r data}
A  <- read.table('../../data/plate_A.tsv', header = TRUE, as.is = TRUE)
B  <- read.table('../../data/plate_B.tsv', header = TRUE, as.is = TRUE)
C  <- read.table('../../data/plate_C.tsv', header = TRUE, as.is = TRUE)
D  <- read.table('../../data/plate_D.tsv', header = TRUE, as.is = TRUE)
i7 <- read.table('../../data/i7_indices.tsv', col.names = c('id', 'inAdapter', 'inSheet'), row.names = 1, as.is = TRUE)
i5 <- read.table('../../data/i5_indices.tsv', col.names = c('id', 'inAdapter', 'inSheet1', 'inSheet2'), row.names = 1, as.is = TRUE)
stopifnot(all.equal(rownames(A), rownames(B)))
stopifnot(all.equal(rownames(C), rownames(D)))
stopifnot(all.equal(colnames(A), colnames(C)))
stopifnot(all.equal(colnames(B), colnames(D)))

AllSamples <- as.matrix(rbind(cbind(A,B), cbind(C,D)))

IndexSample <- do.call(
  rbind,
  lapply(colnames(AllSamples), function(x) {
    t(sapply(rownames(AllSamples), function(y, i=x) c(AllSamples[y, i], i, i7[i, 'inAdapter'], y, i5[y, 'inAdapter']), USE.NAMES = FALSE))
  })
)

colnames(IndexSample) <- c('Sample','i7','i7seq','i5','i5seq')
IndexSample <- as.data.frame(IndexSample)
IndexSample$Sample <- as.character(IndexSample$Sample)
RealSamples <- ! is.na(IndexSample$Sample)

stopifnot(sum(startsWith(IndexSample$Sample, 'HOP'), na.rm = TRUE) == 0)

IndexSample[is.na(IndexSample$Sample), 'Sample'] <- sprintf("HOP%03i", 1:sum(is.na(IndexSample$Sample)))

IndexSample <- IndexSample[! is.na(IndexSample[,'Sample']),]
write.table(IndexSample, file = 'SampleIndexSeq.tsv', sep = '\t', row.names = FALSE)

kable(IndexSample)
```

The table above can be used to fill up the sample sheet for the sequencing service. Because samples
are tagged with Illumina indices, rather than inline barcodes, we are at the machine's mercy
when it comes to demultiplexing. Would I be able to know what or how many reads were attributed
to a particular combination of indices without a real sample assigned to it? Probably not, unless
I tell the machine that such combination is actually assigned to a (*virtual*) sample. For this reason,
I include in the data sheet all actually not used combinations as assigned to as many non-existent, virtual
samples the names of which start with *HOP*.

Below I check the Hamming distances among them, using
an algorithm suggested by [Zheyuan Li](https://stackoverflow.com/users/4891738/zheyuan-li).

```{r hamming}
# The following is adapted from https://stackoverflow.com/a/51918595
i7.used <- unique(as.character(IndexSample[RealSamples,'i7seq']))
MinHamming.i7 <- do.call(
  rbind,
  lapply(i7.used, function(x, y = i7.used){
    A <- strsplit(x, split = '')[[1]]
    B <- do.call(cbind, strsplit(y, split = ''))
    H <- colSums(A != B)
    minH <- min(H[H > 0])
    whichMinH <- which(H == minH)
    whatMinH  <- paste(y[whichMinH], collapse = ',')
    return(data.frame(Index = x, MinHamming = minH, Closest = whatMinH))
  })
)
kable(MinHamming.i7)

i5.used <- unique(as.character(IndexSample[RealSamples, 'i5seq']))
MinHamming.i5 <- do.call(
  rbind,
  lapply(i5.used, function(x, y = i5.used){
    A <- strsplit(x, split = '')[[1]]
    B <- do.call(cbind, strsplit(y, split = ''))
    H <- colSums(A != B)
    minH <- min(H[H > 0])
    whichMinH <- which(H == minH)
    whatMinH  <- paste(y[whichMinH], collapse = ',')
    return(data.frame(Index = x, MinHamming = minH, Closest = whatMinH))
  })
)
kable(MinHamming.i5)
```

Index S517 is at Hamming distance of at least 5 from any other. Thus, it's the only
one that can be read with up to 2 errors.

## Sesion Info

```{r sessionInfo}
sessionInfo()
```

