---
title: "K-mer analysis"
author: "J. Ignacio Lucas Lledó"
date: "26/11/2021"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I want to explore some aspects of the first batch of fastq files received
using the K-mer Analysis Toolkit [@Mapleson2016]. K-mer counts offer a signature
of the sequence compostion of a fastq file. The distribution of counts is related
to how repetitive the content is. Many k-mers (words) counted few times indicate
that most reads are quite different among them, as if coverage was very low or
if they had a lot of random errors.

We can also compare the counts between two fastq files: the more similar the
sequences in the two files, the more correlated the counts will be. With this,
I can take a look at how much more similar missassigned reads are to those from
their putative original sample than to those from any other sample. I should
also be able to confirm that samples from the same species have more similar
sequence compositions than unrelated fastq files.

Finally, I noticed that fastq files have reads from two lanes (indicated in
the fourth ":"-separated field of the name of the read), and we have been
told that lane 2 delivered low quality data. If I manage to distribute a fastq
file in two, for the two lanes, I could also compare them.

K-mers can be counted on the sequenced strand or on both. Reads are oriented,
because specific adapters were ligated to either enzyme's cut site. That is,
reads from a locus will always be sequenced from the same side. Thus, I don't
need to count *canonical* k-mers, but only in one strand.

## General k-mer profiles

I choose a few fastq files to analyse, just as examples.

1. One per plate, from different species: LAN039, BRZ057, THU185 and WAL078.
2. One of the "control" samples with a unique dual index: 175619.
3. Four mock samples sharing an index with the control sample: HOP006, HOP026,
HOP034 and HOP49.
4. And three mock samples from another sector of plate D: HOP038, HOP052 and HOP066.

```{bash}
CHOSEN=(LAN039 BRZ057 THU185 WAL078 175619 HOP006
        HOP049 HOP026 HOP034 HOP052 HOP066 HOP038)

for sample in ${CHOSEN[@]}; do
   if [ ! -d $sample ]; then mkdir $sample; fi
   for fastq in $(ls ../../data/fastq/*.fastq.gz | grep $sample); do
      READ=$(echo $fastq | cut -d '_' -f 4)
      if [ ! -e $sample/$READ.hist ]; then
         kat hist -t 2 -o -N $sample/$READ $fastq &> $sample/$READ.log &
      fi
   done
done
wait

if [ ! -e byRead ]; then
   gawk '(FNR == 1){
      split(FILENAME, A, /\//)
      SAMPLE = A[2]
      READ = substr(A[3], 2, 1)
      ALLSAMPLES[SAMPLE] = 1
   }(/^[^#]/){
      F[SAMPLE, READ, $1] = $2
   }END{
      for (s in ALLSAMPLES) {
         for (i = 1; i <= 2; i++) {
            for (j = 1; j <= 2000; j++) {
               print s "\t" i "\t" j "\t" F[s, i, j]
            }
         }
      }
   }' $(find . -name 'R[12]') | grep -v HOP > byRead.hist
fi
```

```{r plotReads}
library('ggplot2')
byRead <- read.table('byRead.hist', col.names = c('Sample','Read','kmerFreq','kmerNum'),
                     colClasses = c('factor','factor','numeric','numeric'))
ggplot(data = byRead, mapping = aes(x = kmerFreq, y = kmerNum, color = Read)) +
   geom_line() + xlim(1, 500) + scale_y_log10() + facet_wrap(~Sample)
```

Actaully, the mock samples do not have enough reads for a meaningful k-mer profile.
In the rest, we can see the distribution of k-mer frequencies does resemble a mixture
of an exponential (sequencing errors and possible contamination) and a normal that
could be centered around 100. The problem is that the excess of low frequency k-mers
makes it difficult to see the normal component. Changing the *k* up to 51 does not
improve the profile.

Importantly, forward and reverse reads have very similar profiles.

## Comparison between lane 1 and lane 2

The challenge is to split fastq files by lane. I can use *gawk*. In
ordre to avoid writing and reading from additional fastq files, I wanted
to make `kat hist` read from standard input, but it does not work.
I raised [an issue](https://github.com/TGAC/KAT/issues/163) in KAT's
github. For the moment, I will have to use files.

```{bash lanes}
CHOSEN=(LAN039 BRZ057 THU185 WAL078 175619)
for sample in ${CHOSEN[@]}; do
   if [ ! -e $sample/lane1.hist ]; then
      if [ ! -e $sample/lane1.fasta ]; then
         gunzip -c ../../data/fastq/$sample*.fastq.gz | \
         gawk -v SAMPLE=$sample '(NR % 4 == 1){
            if (NR > 1) {
               split(NAME,A,/:/)
               printf ">%s\n%s\n", NAME, SEQ >> sprintf("%s/lane%i.fasta", SAMPLE, A[4])
            }
            NAME = $1
         }(NR % 4 == 2){
            SEQ  = $1
         }END{
            split(NAME, A, /:/)
            printf "%s\n%s\n", NAME, SEQ >> sprintf("%s/lane%i.fasta", SAMPLE, A[4])
         }' &
      fi
   fi
done
wait

for sample in ${CHOSEN[@]}; do
   if [ ! -e $sample/lane1.hist ]; then
      kat hist -N -t 4 -h 2000 -o $sample/lane1.hist $sample/lane1.fasta &> $sample/lane1.log &
      kat hist -N -t 4 -h 2000 -o $sample/lane2.hist $sample/lane2.fasta &> $sample/lane2.log &
   fi
done
wait
if [ ! -e byLane.hist ]; then
   gawk '(FNR == 1){
      split(FILENAME, A, /\//)
      SAMPLE = A[2]
      LANE = substr(A[3], 5, 1)
      ALLSAMPLES[SAMPLE] = 1
   }(/^[^#]/){
      F[SAMPLE, LANE, $1] = $2
   }END{
      for (s in ALLSAMPLES) {
         for (i = 1; i <= 2; i++) {
            for (j = 1; j <= 2000; j++) {
               print s "\t" i "\t" j "\t" F[s, i, j]
            }
         }
      }
   }' $(find . -name 'lane*.hist') > byLane.hist
fi
```

```{r plotLanes}
byLane <- read.table('byLane.hist', col.names=c('Sample','Lane','kmerFreq','kmerNum'),
                     colClasses = c('factor','factor','numeric','numeric'))
ggplot(data = byLane, mapping = aes(x = kmerFreq, y = kmerNum, color = Lane)) +
   geom_line() + scale_y_log10() + xlim(1, 500) + facet_wrap(~Sample)
```

Lanes 1 and 2 have almost identical k-mer profiles in all 5 samples
analysed here. Filtering low quality reads could improve the profile,
if sequencing errors are a big portion of the low multiplicity peak
in the k-mer profile. But contamination and sequencing of fragments
from out of the selected range can also contribute that that peak.

## K-mer and GC content

```{bash}
CHOSEN=(LAN039 BRZ057 THU185 WAL078 175619)

for sample in ${CHOSEN[@]}; do
   if [ ! -e $sample/gcp.mx ]; then
      kat gcp -N -t 4 -o $sample/gcp ../../data/fastq/$sample*.fastq.gz &> $sample/gcp.log &
   fi
done
wait
```

```{r plotGC, fig.width=20, fig.height=20, warning=FALSE}
library('tidyr')
library('gridExtra')
for (Sample in c('LAN039', 'BRZ057', 'THU185', 'WAL078', '175619')) {
   gcp      <- read.table(paste(Sample, 'gcp.mx', sep='/'))
   gcp$GC   <- seq(0,26)
   
   gcp_long <- pivot_longer(gcp, 1:1001, names_to = 'kmerFreq', values_to = 'kmerNum')
   gcp_long$kmerFreq <- as.numeric(substring(gcp_long$kmerFreq, 2)) - 1
   assign(sprintf('p%s', Sample), ggplot(data=gcp_long, mapping=aes(x=kmerFreq, y=GC, fill=kmerNum)) +
      scale_fill_gradientn(trans='log', colors=terrain.colors(10), breaks=c(1,100,10000,1e6), na.value='blue') +
      geom_tile() + ggtitle(Sample))
}

grid.arrange(pLAN039, pBRZ057, pTHU185, pWAL078, p175619, nrow=3)
```


## Comparisons between samples

```{bash comp}
SAMPLE=(LAN039 BRZ057 THU185 WAL078 175619)
for first in 0 1 2 3; do
   for second in $(seq $(($first + 1)) 4); do
      if [ ! -e ${SAMPLE[$first]}/comp_${SAMPLE[$second]}-main.mx ]; then
         FASTQ1=$(ls -1 ../../data/fastq/${SAMPLE[$first]}* | grep R1)
         FASTQ2=$(ls -1 ../../data/fastq/${SAMPLE[$second]}* | grep R1)
         if [ -z "${SIZE[$second]}" ]; then
            SIZE[$second]=$(gunzip -c $FASTQ2 | wc -l)
         fi
         if [ -z "${SIZE[$first]}" ]; then
            SIZE[$first]=$(gunzip -c $FASTQ1 | wc -l)
         fi
         if [ ${SIZE[$first]} -ge ${SIZE[$second]} ]; then
            PARAM='--d2_scale'
            VALUE=$(echo "${SIZE[$second]} / ${SIZE[$first]}" | bc -l)
         else
            PARAM='--d1_scale'
            VALUE=$(echo "${SIZE[$first]} / ${SIZE[$second]}" | bc -l)
         fi
         kat comp --output_prefix ${SAMPLE[$first]}/comp_${SAMPLE[$second]} \
                  --non_canonical_1 \
                  --non_canonical_2 \
                  --threads 8 \
                  $PARAM $VALUE \
                  --density_plot \
                  $FASTQ1 $FASTQ2 \
                  &> ${SAMPLE[$first]}/comp_${SAMPLE[$second]}.log &
      fi
   done
done
wait
```

```{r plotComp, fig.width=20, fig.height=20, warning=FALSE}
SAMPLE = c('LAN039', 'BRZ057', 'THU185', 'WAL078', '175619')
for (i in 1:4) {
   for (j in (i+1):5) {
      comp <- read.table(sprintf('%s/comp_%s-main.mx', SAMPLE[i], SAMPLE[j]))
      comp$Y <- seq(0,1000)
      comp_long <- pivot_longer(comp, 1:1001, names_to = 'X', values_to = 'frequency')
      comp_long$X <- as.numeric(substring(comp_long$X, 2)) - 1
      assign(sprintf('p%s_%s', SAMPLE[i], SAMPLE[j]),
             ggplot(data=comp_long, mapping=aes(x=X, y=Y, fill=frequency)) + geom_tile() + xlim(0,500) + ylim(0,500) +
             scale_fill_gradientn(trans='log', colors=terrain.colors(10), breaks=c(1,100,10000,1000000), na.value='blue') +
             xlab(sprintf('27-mer frequency in %s', SAMPLE[i])) +
             ylab(sprintf('27-mer frequency in %s', SAMPLE[j])))
   }
}

layout_matrix <- matrix(c(1, NA, NA, NA,
                          2,  3, NA, NA,
                          4,  5,  6, NA,
                          7,  8,  9, 10), nrow=4, ncol=4, byrow=TRUE)

grid.arrange(pLAN039_BRZ057,
             pLAN039_THU185, pBRZ057_THU185,
             pLAN039_WAL078, pBRZ057_WAL078, pTHU185_WAL078, 
             pLAN039_175619, pBRZ057_175619, pTHU185_175619, pWAL078_175619,
             layout_matrix = layout_matrix)
```

# References