---
title: "$F_{ST}$ within lakes."
author: "J. Ignacio Lucas Lledó & Mar Llaberia-Robledillo"
date: "`r Sys.Date()`"
output:
   html_document:
      toc: TRUE
bibliography: "../../doc/references.bib"
---

Species in the same lake are supposed to hybridize occasionally, so that they may
share variation along large portions of the genome. That is: the $F_{ST}$ between species
in the same lake are expected to be low, in general. But loci that contribute to the
adaptation of an ecomorph must remain divergent between ecomorphs in the same lake.
In a background of low $F_{ST}$ values, it should be easy to spot the positions with
high $F_{ST}$ values.

# Alpine region 
We will compare first *C. fatioi* (Felchen ecomorph, Thun and Brienz lakes) with
*C. steinmanni* (Balchen ecomorph, in Thun). We use `vcftools` [@Danecek2011] to
estimate weighted $F_{ST}$s [@Weir1984] in overlapping genomic windows of one
million base pairs.

```{bash Thun-Brienz}
VCF=../2022-04-22/assem2_outfiles/assem2.vcf
FISHDATA=../../data/Fish_clean2.tsv

if [ ! -d FST ]; then mkdir FST; fi

if [ ! -e FST/Thun-Brienz.windowed.weir.fst ]; then
   if [ ! -e dirty.recode.vcf ]; then
      vcftools --vcf $VCF \
               --out dirty \
               --remove ../2022-05-04/TheWorst.txt \
               --recode \
               --recode-INFO-all > vcftools.log 2>&1 
   fi

   if [ ! -e C.fatioi.txt ]; then
      gawk '($19 == "C.fatioi"){print $1}' $FISHDATA > C.fatioi.txt
   fi

   if [ ! -e C.steinmanni ]; then
      gawk '($19 == "C.steinmanni"){print $1}' $FISHDATA > C.steinmanni.txt
   fi

   vcftools --vcf dirty.recode.vcf \
            --out FST/Thun-Brienz \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop C.fatioi.txt \
            --weir-fst-pop C.steinmanni.txt \
            --fst-window-size 1000000 \
            --fst-window-step  250000 >> vcftools.log 2>&1
fi
```

```{r plots1}
suppressMessages(library(Biostrings))
suppressMessages(library(GenomicRanges))
suppressMessages(library(ggplot2))
ref             <- readDNAStringSet('../../data/reference.fa')
# The names of chromosomes must be only the first word in the fasta name line.
names(ref)      <- sapply(strsplit(names(ref), " "), '[[', 1)
# the refWidth table contains the lengths of chromosomes.
refWidth        <- width(ref)
names(refWidth) <- names(ref)

fst1 <- read.table('FST/Thun-Brienz.windowed.weir.fst',
                  header = TRUE, stringsAsFactors = FALSE)
fst1$POS        <- (fst1$BIN_END + fst1$BIN_START) / 2
offset          <- cumsum(refWidth)[1:39]
names(offset)   <- names(ref)[2:40]
offset['LR778253.1'] <- 0
fst1$offset      <- offset[fst1$CHROM]
fst1$POS2        <- fst1$POS + fst1$offset
ChromCol        <- rep(c(1,2), 20)
names(ChromCol) <- names(ref)
fst1$ChromCol    <- ChromCol[fst1$CHROM]

# Thun-Brienz:
thrs1 <- quantile(fst1$WEIGHTED_FST, probs = 0.96)
f1 <- fst1$WEIGHTED_FST >= thrs1

ggplot(data = fst1, mapping = aes(x = POS2, y = WEIGHTED_FST, color = ChromCol)) +
   geom_point() + theme(legend.position="none") +
   xlab('Position') + ylab(expression(F[ST])) +
   ggtitle('Felchen vs. Balchen ecomoprhs (Thun-Brienz lakes)') +
   geom_hline(yintercept = round(thrs1, 2), color = "red") +
   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Above, $F_{ST}$ values are plotted against their positions in the genome,
with consecutive chromosomes in alternative colors. Every data point is a
weighted $F_{ST}$ value among a few SNPs in a 1000000-bp genomic window.
Large overlaps between adjacent windows would make a smoother line in
individual chromosomes, but give the false impression of a lot of data.

Some $F_{ST}$ values are negative, but close to zero. Negative $F_{ST}$ values
must mean that individuals from the same species are less similar among them
than to members of the other species, at least in these samples. I assume the low
magnitude of negative values is compatible with random sampling variation.

The equivalent comparison between a Felchen and a Balchen ecomorph in a different
lake is the comparsion of *C. zuerichensis* (Felchen) and *C. duplex* (Balchen),
both from the Zurich and Walen lakes. If natural selection targeted the same loci
in both lakes to produce the adaptations that define the Felchen and Balchen
ecomorphs, then, several windows with a high $F_{ST}$ above should also have a
high $F_{ST}$ in the results below. Let's see.

```{bash zuerichensis-duplex}
FISHDATA=../../data/Fish_clean2.tsv

if [ ! -e FST/Walen-Zurich.windowed.weir.fst ]; then
   if [ ! -e C.zuerichensis.txt ]; then
      gawk '($19 == "C.zuerichensis"){print $1}' $FISHDATA > C.zuerichensis.txt
   fi
   
   if [ ! -e C.duplex.txt ]; then
      gawk '($19 == "C.duplex"){print $1}' $FISHDATA > C.duplex.txt
   fi
   vcftools --vcf dirty.recode.vcf \
            --out FST/Walen-Zurich \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop C.zuerichensis.txt \
            --weir-fst-pop C.duplex.txt \
            --fst-window-size 1000000 \
            --fst-window-step  250000 >> vcftools.log 2>&1
fi
```


```{r plots2}

fst2 <- read.table('FST/Walen-Zurich.windowed.weir.fst',
                   header = TRUE, stringsAsFactors = FALSE)
fst2$POS        <- (fst2$BIN_END + fst2$BIN_START) / 2
offset          <- cumsum(refWidth)[1:39]
names(offset)   <- names(ref)[2:40]
offset['LR778253.1'] <- 0
fst2$offset     <- offset[fst2$CHROM]
fst2$POS2       <- fst2$POS + fst2$offset
ChromCol        <- rep(c(1,2), 20)
names(ChromCol) <- names(ref)
fst2$ChromCol   <- ChromCol[fst2$CHROM]

# Walen-Zurich:
thrs2 <- quantile(fst2$WEIGHTED_FST, probs = 0.96)
f2 <- fst2$WEIGHTED_FST >= thrs2

ggplot(data = fst2, mapping = aes(x = POS2, y = WEIGHTED_FST, color = ChromCol)) +
   geom_point() + theme(legend.position="none") +
   xlab('Position') + ylab(expression(F[ST])) +
   ggtitle('Felchen vs. Balchen ecomorphs (Walen-Zurich lakes)') +
   geom_hline(yintercept = round(thrs2, 2), color = "red") +
   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

In lakes Zurich and Walen, which are connected, $F_{ST}$ values between the
Balchen and Felchen ecomorphs are higher in general, indicating that the
species *C. zuerichensis* and *C. duplex* are better differentiated genetically
than *C. fatioi* and *C. steinmanni* in Thun-Brienz.

We assume that the highest values of $F_{ST}$ between two species in the same
lake must be associated with loci adapted to each species' environment. However,
we know that $F_{ST}$ is a *relative* measure of divergence between populations,
because it is influenced by the levels of diversity within populations [@Cruickshank2014].
That is: the lower the genetic diversity within populations, the higher $F_{ST}$
is expected to be. We will proceed to identify and compare windows of high
differentiation, and then we will check if they could be an artifact of
variable levels of within-population diversity.

The first way to compare highly differentiated genomic regions between lakes
consists in: defining a threshold and visualizing the positions of windows
above that threshold in both lakes. The threshold is the 96$^\text{th}$ percentile.

```{r comparacio1, fig.width=10, fig.height=3}
# Thun-Brienz:
thrs1 <- quantile(fst1$WEIGHTED_FST, probs = 0.96)
f1 <- fst1$WEIGHTED_FST >= thrs1
# Walen-Zurich:
thrs2 <- quantile(fst2$WEIGHTED_FST, probs = 0.96)
f2 <- fst2$WEIGHTED_FST >= thrs2

GR1 <- GRanges(
   seqnames = factor(fst1[f1, 'CHROM']),
   ranges = IRanges(start = fst1[f1, 'BIN_START'],
                    end = fst1[f1, 'BIN_END'],
                    names = paste0(fst1[f1, 'CHROM'], ':',
                                   fst1[f1, 'BIN_START'], '-',
                                   fst1[f1, 'BIN_END'],
                                   sep = '')),
   strand = '+',
   seqinfo = Seqinfo(seqnames = names(ref),
                     seqlengths = width(ref))
)
GR1 <- trim(GR1)
GR1 <- reduce(GR1)

GR2 <- GRanges(
   seqnames = factor(fst2[f2, 'CHROM']),
   ranges = IRanges(start = fst2[f2, 'BIN_START'],
                    end = fst2[f2, 'BIN_END'],
                    names = paste0(fst2[f2, 'CHROM'], ':',
                                   fst2[f2, 'BIN_START'], '-',
                                   fst2[f2, 'BIN_END'],
                                   sep = '')),
   strand = '+',
   seqinfo = Seqinfo(seqnames = names(ref),
                     seqlengths = width(ref))
)
GR2 <- trim(GR2)
GR2 <- reduce(GR2)

overlaps <- findOverlaps(GR1, GR2, maxgap = 50000)
overlaps
AlpOverlaps <- reduce(c(GR1[queryHits(overlaps)],
                        GR2[subjectHits(overlaps)]))

GR <- data.frame(
   chr = c(seqnames(GR1), seqnames(GR2)),
   start = c(start(GR1), start(GR2)),
   end = c(end(GR1), end(GR2)),
   lake = c(rep('Thun-Brienz', length(GR1)),
            rep('Walen-Zurich', length(GR2))),
   height = c(rep(2, length(GR1)),
              rep(1, length(GR2)))
)

for (chr in unique(GR$chr)) {
   f <- GR$chr == chr
   p <- ggplot(data = GR[f,],
               mapping = aes(x = start,
                             y = height,
                             xend = end,
                             yend = height,
                             color = lake)) +
      geom_segment(size = 5) +
      ylim(0,3) +
      xlim(0, width(ref[chr])) +
      theme(aspect.ratio = 0.3,
            axis.text.y = element_blank(),
            axis.line.y = element_blank(),
            axis.title.y = element_blank(),
            axis.ticks.y = element_blank()) +
      ggtitle(chr) + xlab('Position (bp)')
   print(p)
}
```

Only `r length(overlaps)` genomic regions are highly differentiated
between ecomorphs in both Alpine lake systems.

# Arctic region

First, I will compare P and L ecomorphs from Suopatjavri. I will
recover of the text files created in the previous folder (2022-12-23)
for the estimation.

```{bash P-L-SUO}
if [ ! -e FST/Suo.windowed.weir.fst ]; then
   P=../2022-12-23/SUO_P.txt
   L=../2022-12-23/SUO_L.txt
   
   vcftools --vcf dirty.recode.vcf \
            --out FST/Suo \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop $P \
            --weir-fst-pop $L \
            --fst-window-size 1000000 \
            --fst-window-step  250000 >> vcftools.log 2>&1
fi
```


```{r plots3}
fst3 <- read.table('FST/Suo.windowed.weir.fst',
                  header = TRUE, stringsAsFactors = FALSE)
fst3$POS    <- (fst3$BIN_END + fst3$BIN_START) / 2
offset     <- cumsum(refWidth)[1:39]
names(offset) <- names(ref)[2:40]
offset['LR778253.1'] <- 0
fst3$offset <- offset[fst3$CHROM]
fst3$POS2 <- fst3$POS + fst3$offset
ChromCol <- rep(c(1,2), 20)
names(ChromCol) <- names(ref)
fst3$ChromCol <- ChromCol[fst3$CHROM]

# SUO:
thrs3 <- quantile(fst3$WEIGHTED_FST, probs = 0.96)
f3 <- fst3$WEIGHTED_FST >= thrs3

ggplot(data = fst3, mapping = aes(x = POS2, y = WEIGHTED_FST, color = ChromCol)) +
   geom_point() + theme(legend.position="none") +
   xlab('Position') + ylab(expression(F[ST])) +
   ggtitle('Profundal vs. Littoral ecomorphs (Suopatjavri)') +
   geom_hline(yintercept = round(thrs3, 2), color = "red") +
   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Second, I'm goingo to compare P and L from Langfjordvatn. The same as above. 

```{bash P-L-LAN}
if [ ! -e FST/Lan.windowed.weir.fst ]; then
   P=../2022-12-23/LAN_P.txt
   L=../2022-12-23/LAN_L.txt
   
   vcftools --vcf dirty.recode.vcf \
            --out FST/Lan \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop $P \
            --weir-fst-pop $L \
            --fst-window-size 1000000 \
            --fst-window-step  250000 >> vcftools.log 2>&1
fi
```


```{r plots4}
fst4 <- read.table('FST/Lan.windowed.weir.fst',
                  header = TRUE, stringsAsFactors = FALSE)
fst4$POS    <- (fst4$BIN_END + fst4$BIN_START) / 2
offset     <- cumsum(refWidth)[1:39]
names(offset) <- names(ref)[2:40]
offset['LR778253.1'] <- 0
fst4$offset <- offset[fst4$CHROM]
fst4$POS2 <- fst4$POS + fst4$offset
ChromCol <- rep(c(1,2), 20)
names(ChromCol) <- names(ref)
fst4$ChromCol <- ChromCol[fst4$CHROM]

# LAN:
thrs4 <- quantile(fst4$WEIGHTED_FST, probs = 0.96)
f4 <- fst4$WEIGHTED_FST >= thrs4

ggplot(data = fst4, mapping = aes(x = POS2, y = WEIGHTED_FST, color = ChromCol)) +
   geom_point() + theme(legend.position="none") +
   xlab('Position') + ylab(expression(F[ST])) +
   ggtitle('Profundal vs. Littoral ecomorphs (Langfjordvatn)') +
   geom_hline(yintercept = round(thrs4, 2), color = "red") +
   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

```

```{r comparacio2, fig.width=10, fig.height=3}
# SUO:
thrs3 <- quantile(fst3$WEIGHTED_FST, probs = 0.96)
f3 <- fst3$WEIGHTED_FST >= thrs3
# LAN:
thrs4 <- quantile(fst4$WEIGHTED_FST, probs = 0.96)
f4 <- fst4$WEIGHTED_FST >= thrs4

GR3 <- GRanges(
   seqnames = factor(fst3[f3, 'CHROM']),
   ranges = IRanges(start = fst3[f3, 'BIN_START'],
                    end = fst3[f3, 'BIN_END'],
                    names = paste0(fst3[f3, 'CHROM'], ':',
                                   fst3[f3, 'BIN_START'], '-',
                                   fst3[f3, 'BIN_END'],
                                   sep = '')),
   strand = '+',
   seqinfo = Seqinfo(seqnames = names(ref),
                     seqlengths = width(ref))
)
GR3 <- trim(GR3)
GR3 <- reduce(GR3)

GR4 <- GRanges(
   seqnames = factor(fst4[f4, 'CHROM']),
   ranges = IRanges(start = fst4[f4, 'BIN_START'],
                    end = fst4[f4, 'BIN_END'],
                    names = paste0(fst4[f4, 'CHROM'], ':',
                                   fst4[f4, 'BIN_START'], '-',
                                   fst4[f4, 'BIN_END'],
                                   sep = '')),
   strand = '+',
   seqinfo = Seqinfo(seqnames = names(ref),
                     seqlengths = width(ref))
)
# vcftools creates windows out of the end of the chromosome.
GR4 <- trim(GR4)
GR4 <- reduce(GR4)

overlaps <- findOverlaps(GR3, GR4, maxgap = 10000)
overlaps
ArcOverlaps <- reduce(c(GR3[queryHits(overlaps)],
                        GR4[subjectHits(overlaps)]))
ArcOverlaps

GRn <- data.frame(
   chr = c(seqnames(GR3), seqnames(GR4)),
   start = c(start(GR3), start(GR4)),
   end = c(end(GR3), end(GR4)),
   lake = c(rep('Suopatjavri', length(GR3)),
            rep('Langfjordvatn', length(GR4))),
   height = c(rep(1, length(GR3)),
              rep(2, length(GR4)))
)

for (chr in unique(GRn$chr)) {
   f <- GRn$chr == chr
   p <- ggplot(data = GRn[f,],
               mapping = aes(x = start,
                             y = lake,
                             xend = end,
                             yend = lake,
                             color = lake)) +
      geom_segment(size = 5) +
      #ylim(0,3) +
      xlim(0, width(ref[chr])) +
            theme(aspect.ratio = 0.3,
            axis.text.y = element_blank(),
            axis.line.y = element_blank(),
            axis.title.y = element_blank(),
            axis.ticks.y = element_blank()) +
      ggtitle(chr) + xlab('Position (bp)')
   print(p)
}
```

It turns out that a few genomic regions are highly differentiated between
ecomorphs simultaneously in the two Arctic lakes. Now, the question is if
any of the `r length(overlaps)` regions presumably important in ecological
adaptation in the Arctic are also important for the adaptation in the Alpine
region.

```{r crossRegion, fig.width=10, fig.height=3}
crossRegionOverlaps <- findOverlaps(AlpOverlaps,
                                    ArcOverlaps,
                                    maxgap = 10000)
crossRegionOverlaps
chromHighFst <- as.character(
   unique(seqnames(AlpOverlaps[queryHits(crossRegionOverlaps)]))
)
fst1$lake <- 'Thun-Brienz'
fst1$height <- 3
fst2$lake <- 'Walen-Zurich'
fst2$height <- 4
fst3$lake <- 'Suopatjavri'
fst3$height <- 1
fst4$lake <- 'Langfjordvatn'
fst4$height <- 2
AllTables <- rbind(fst1[f1,], fst2[f2,], fst3[f3,], fst4[f4,])
for (chr in chromHighFst) {
   f <- AllTables$CHROM == chr
   p <- ggplot(data = AllTables[f,],
               mapping = aes(x = BIN_START,
                             y = lake,
                             xend = BIN_END,
                             yend = lake,
                             color = WEIGHTED_FST)) +
      geom_segment(size = 5) + xlab('Position (bp)') +
      ggtitle(chr) +
      guides(colour=guide_legend(title=expression(F[ST])))
   print(p)
}
```

There are 5 regions, in 5 chromosomes with approximate overlap among the
4 lakes. These are regions of interest that should be studied further to
see what genes are there.

# Independent windows

For a more formal test of coincidence of highly differentiated regions
between two lakes, we need $F_{ST}$ estimates independent of each other,
that is, from *non-overlaping* windows.

```{bash independence}
if [ ! -e FST/Thun-Brienz.ind.windowed.weir.fst ]; then
   vcftools --vcf dirty.recode.vcf \
            --out FST/Thun-Brienz.ind \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop C.fatioi.txt \
            --weir-fst-pop C.steinmanni.txt \
            --fst-window-size 1000000 \
            --fst-window-step 1000000 >> vcftools.log 2>&1
fi
if [ ! -e FST/Walen-Zurich.ind.windowed.weir.fst ]; then
   vcftools --vcf dirty.recode.vcf \
            --out FST/Walen-Zurich.ind \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop C.zuerichensis.txt \
            --weir-fst-pop C.duplex.txt \
            --fst-window-size 1000000 \
            --fst-window-step 1000000 >> vcftools.log 2>&1
fi

if [ ! -e FST/Suo.ind.windowed.weir.fst ]; then
   vcftools --vcf dirty.recode.vcf \
            --out FST/Suo.ind \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop ../2022-12-23/SUO_P.txt \
            --weir-fst-pop ../2022-12-23/SUO_L.txt \
            --fst-window-size 1000000 \
            --fst-window-step 1000000 >> vcftools.log 2>&1
fi

if [ ! -e FST/Lan.ind.windowed.weir.fst ]; then
   vcftools --vcf dirty.recode.vcf \
            --out FST/Lan.ind \
            --thin 500 \
            --remove-indels \
            --mac 2 \
            --max-alleles 2 \
            --max-missing 1 \
            --max-meanDP 1000 \
            --weir-fst-pop ../2022-12-23/LAN_P.txt \
            --weir-fst-pop ../2022-12-23/LAN_L.txt \
            --fst-window-size 1000000 \
            --fst-window-step 1000000 >> vcftools.log 2>&1
fi
```

```{r newTables}
fst1 <- read.table('FST/Thun-Brienz.ind.windowed.weir.fst',
                   header = TRUE, stringsAsFactors = FALSE)
fst2 <- read.table('FST/Walen-Zurich.ind.windowed.weir.fst',
                   header = TRUE, stringsAsFactors = FALSE)
fst3 <- read.table('FST/Suo.ind.windowed.weir.fst',
                   header = TRUE, stringsAsFactors = FALSE)
fst4 <- read.table('FST/Lan.ind.windowed.weir.fst',
                   header = TRUE, stringsAsFactors = FALSE)
```

# Bivariate distribution of $F_{ST}$s

The plots below give an idea of how few genomic regions are highly
differentiated between ecomorphs simultaneously in two lake systems.
The comparisons between lakes from different regions must be quite
pointless, because we expecte even less coincidence.

```{r correlations}
row.names(fst1) <- paste0(fst1$CHROM, ':', fst1$BIN_START, sep = '')
row.names(fst2) <- paste0(fst2$CHROM, ':', fst2$BIN_START, sep = '')
row.names(fst3) <- paste0(fst3$CHROM, ':', fst3$BIN_START, sep = '')
row.names(fst4) <- paste0(fst4$CHROM, ':', fst4$BIN_START, sep = '')
f <- intersect(intersect(row.names(fst1), row.names(fst2)),
               intersect(row.names(fst3), row.names(fst4)))
commonFst <- data.frame(
   chr    = fst1[f, 'CHROM'],
   start  = fst1[f, 'BIN_START'],
   end    = fst1[f, 'BIN_END'],
   Thu    = fst1[f, 'WEIGHTED_FST'],
   Wal    = fst2[f, 'WEIGHTED_FST'],
   Suo    = fst3[f, 'WEIGHTED_FST'],
   Lan    = fst4[f, 'WEIGHTED_FST']
)

ggplot(commonFst, aes(x=Thu, y=Wal)) + geom_hex() +
   xlab('Lakes Thun and Brienz') +
   ylab('Lakes Walen and Zurich') +
   ggtitle('Alpine region')
ggplot(commonFst, aes(x=Suo, y=Lan)) + geom_hex() +
   xlab('Suopatjavri') + ylab('Langfjordvatn') +
   ggtitle('Arctic region')
```

# Kolmogorov-Smirnof

```{r kolmogorov-smirnof}
thrs1 <- quantile(commonFst$Thu, probs = 0.9)
thrs2 <- quantile(commonFst$Wal, probs = 0.9)
thrs3 <- quantile(commonFst$Suo, probs = 0.9)
thrs4 <- quantile(commonFst$Lan, probs = 0.9)

ks.results <- data.frame(
   chr = unique(commonFst$chr),
   WalThu = NA,
   ThuWal = NA,
   LanSuo = NA,
   SuoLan = NA
)

row.names(ks.results) <- ks.results$chr

for (chr in unique(commonFst$chr)) {
   HighFstThu <- commonFst[commonFst$Thu >= thrs1 & commonFst$chr == chr,]
   HighFstWal <- commonFst[commonFst$Wal >= thrs2 & commonFst$chr == chr,]
   HighFstSuo <- commonFst[commonFst$Suo >= thrs3 & commonFst$chr == chr,]
   HighFstLan <- commonFst[commonFst$Lan >= thrs4 & commonFst$chr == chr,]
   if (dim(HighFstThu)[1] >= 5) {
      ks.results[chr, 'ThuWal'] <- ks.test(HighFstThu$Wal,
                                           commonFst$Wal, exact = FALSE,
                                           alternative = 'greater')$p.value
   }
   if (dim(HighFstWal)[1] >= 5) {
      ks.results[chr, 'WalThu'] <- ks.test(HighFstWal$Thu,
                                           commonFst$Thu, exact = FALSE,
                                           alternative = 'greater')$p.value
   }
   if (dim(HighFstLan)[1] >= 5) {
      ks.results[chr, 'LanSuo'] <- ks.test(HighFstLan$Suo,
                                           commonFst$Suo, exact = FALSE,
                                           alternative = 'greater')$p.value
   }
   if (dim(HighFstSuo)[1] >= 5) {
      ks.results[chr, 'SuoLan'] <- ks.test(HighFstSuo$Lan,
                                           commonFst$Lan, exact = FALSE,
                                           alternative = 'greater')$p.value
   }
}
```

# Session Information

```{r sessionInfo}
sessionInfo()
```

# References