---
title: "Preparing for ipyrad"
author: "J. Ignacio Lucas Lledó"
date: "21/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are three issues we need to solve before using `ipyrad`.

- Fastq file names are too long.
- We have fastq files from two sequencing runs.
- We need to call ipyrad, in its conda environment, from RStudio.

## The fastq file names
Sample names are inferred from the fastq file names, which include:
the sample name, a misterious codeword, and the sequences of the two
indices, i5 and i7. We can remove all that redundant information.
The solution that preserves the data better is to create symbolic
links.

```{bash symlinks}
FASTQDIR1=../../data/fastq
FASTQDIR2=../../data/fastr
DATADIR=/gata/joiglu/cophyhopa/data
if [ ! -d run1 ]; then mkdir run1; fi
if [ ! -d run2 ]; then mkdir run2; fi

# ls -1 /gata/joiglu/cophyhopa/data/fastq/*.fastq.gz | head -n 5
# /gata/joiglu/cophyhopa/data/fastq/175600_12752AAC_GTAGAGGA-CTTAATAG_R1_001.fastq.gz
# /gata/joiglu/cophyhopa/data/fastq/175600_12752AAC_GTAGAGGA-CTTAATAG_R2_001.fastq.gz
# /gata/joiglu/cophyhopa/data/fastq/175601_12765AAC_GCTCATGA-AGCTAGAA_R1_001.fastq.gz
# /gata/joiglu/cophyhopa/data/fastq/175601_12765AAC_GCTCATGA-AGCTAGAA_R2_001.fastq.gz
# /gata/joiglu/cophyhopa/data/fastq/175602_12767AAC_GCTCATGA-TCTTACGC_R1_001.fastq.gz

# First sequencing run:
for file in $(ls -1 $DATADIR/fastq/*.fastq.gz); do
   SAMPLE=$(echo $file | cut -d "/" -f 7 | cut -d "_" -f 1)
   READ=$(echo $file | cut -d "_" -f 4)
   if [ ! -e run1/${SAMPLE}_${READ}.fastq.gz ]; then
      ln -s $file run1/${SAMPLE}_${READ}.fastq.gz
   fi
done

# Now for the second sequencing run:
for file in $(ls -1 $DATADIR/fastr/*.fastq.gz); do
   SAMPLE=$(echo $file | cut -d "/" -f 7 | cut -d "_" -f 1)
   READ=$(echo $file | cut -d "_" -f 4)
   if [ ! -e run2/${SAMPLE}_${READ}.fastq.gz ]; then
      ln -s $file run2/${SAMPLE}_${READ}.fastq.gz
   fi
done
```

## How to activate ipyrad

```{r reticulate}
library('reticulate')
use_condaenv('ipyrad')
```

That's it. Now ipyrad should be available. I will save a copy of the ipyrad
environment, for reproducibility

```{bash exportEnv}
/opt/miniconda3/bin/conda list -n ipyrad --explicit > spec-ipyrad.txt
```

## The two sequencing runs
One good thing of ipyrad is that we can start working with two sets of
fastq files and then tell ipyrad to merge them. As long as the files are
correctly named, it will correctly join the information from files from
the same samples. Thus, we don't need to create new fastq files with the
combined information from the two runs for every sample.

The procedure is described in ipyrad manual as [one library and multiple lanes of sequencing](https://ipyrad.readthedocs.io/en/master/8-branching.html#one-library-multiple-lanes-of-sequencing).

