---
title: "ddRAD protocol"
author: "J. Ignacio Lucas Lledó"
date: "08/06/2021"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Annealing of adapters
Adapter P1 will bind *HindIII* ends, and adapter P2 will bind *SphI* ends. Both adapters need to be annealed separately. Oligos are asumed to be suspended to 100 $/mu$M concentrations. The annealing buffer stock at 10$\times$ concentration should be 100 mM Tris HCl (pH 8), 500 mM NaCl, 10 mM EDTA [@Peterson2012].

```{r AnnealingBufferStock, echo=FALSE}
library(knitr)
AnnealingBufferStock <- data.frame(
  Reagent = c('Tris-HCl 1M pH 8.0', 'EDTA 0.5M', 'NaCl 1M', 'Water'),
  Amount = c(20, 4, 100, 76),
  Final = c(100, 10, 500, NA)
)
kable(AnnealingBufferStock, caption='10$\\times$ Annealing Buffer Stock, 200 $\\mu$l',
      col.names=c('Reagent', 'Amount to add ($\\mu$l)', 'Final concentration (mM)'))
```

Both adapters can be at a final concentration of 2.5 $\mu$M. We will use 1 $\mu$l per sample. We can prepare 500 $\mu$l of annealed stock of each adapter, to have them in a comfortable excess.

- **Adapter P1**: Mix 12.5 $\mu$l *HindIII*-P1.1 100 $\mu$M and 12.5 $\mu$l of *HindIII*-P1.2 100 $\mu$M oligos, 50 $\mu$l of 10$\times$ annealing buffer stock (see above), and 425 $\mu$l of water.
- **Adapter P2**: Mix 12.5 $\mu$l *SphI*-P2.1 100 $\mu$M and 12.5 $\mu$l of *SphI*-P2.2 100 $\mu$M oligos, 50 $\mu$l of 10$\times$ annealing buffer stock (see above), and 425 $\mu$l of water.

# Double digestion
@Salas-Lizana2018 set up digestion reactions in small volumes (9 $\mu$l), where DNA constitutes 67\% of the volume (in there in TE, which contain 1 mM Na$^+$ and 10 mM Cl$^-$, I think). On top of that, they add 0.45 $\mu$l of 1 M NaCl, reaching a salt concentration above 150 mM. That is higher than the salt concentration of a NEBuffer 3.1 (100 mM NaCl), where *SphI*-HF and *HindIII*-HF have 10% activity. That is, our enzymes are much more sensitive to salt than those used by @Salas-Lizana2018. Thus, we need to increase the volume of the digestion reaction to prevent salt-inhibition of the enzymes. The recommendation is to keep DNA volume below 25% of the total volume.

I assume enzymes are at 20000 U/ml. In general, 10 U of enzyme (1 $\mu$l) are supposed to be enough for 1 $\mu$g of DNA. We are using up to 300 ng of DNA in the reaction. If the number of units required are proportional to the amount of DNA, 0.3 $\mu$l of each enzyme (3 U) should be enough for 300 ng. We will use 0.5 $\mu$l.

```{r digestion, echo=FALSE}
DNA.volume <- 6 # µl
TotalReactionVolume <- DNA.volume * 4 # µl
Buffer10XVolume <- 0.1 * TotalReactionVolume
SphI.volume <- 0.5
HindIII.volume <- 0.5
Water.volume <- TotalReactionVolume - Buffer10XVolume - DNA.volume - SphI.volume - HindIII.volume
DigestionMix <- data.frame(
  Component = c('10$\\times$ rCutSmart Buffer', 'SphI-HF', 'HindIII-HF', 'Nuclease-free water'),
  Volum = c(Buffer10XVolume, SphI.volume, HindIII.volume, Water.volume)
)
kable(DigestionMix, caption='Double digestion master mix. Multiply volumes by 1.2 times the number of samples. The total volume of master mix per sample is 18 $\\mu$l',
      col.names = c('Component', 'Volum per reaction ($\\mu$l)'))
```

1. Prepare the double digestion master mix (see table above) for 1.2$\times$ the number of samples.
2. Mix `r TotalReactionVolume - DNA.volume` $\mu$l of double digestion master mix with 6 $\mu$l of sample DNA in each PCR tube or well.
3. Centrifuge briefly (spin down?) and incubate at 37$^\circ{}$C for 4 hours in a thermal cycler with a heated lid. Note: both enzymes are "Time-Saver qualified" (according to NEB), so they are supposed to do the job in 15 minutes. Worth trying?
4. Do **not** heat inactivate the enzymes, because adapters are designed not to reconstitute the restriction target sites upon ligation.

# Ligation of adapters


# Clean up with magnetic beads

# PCR amplification with incorporation of Nextera indices

# Pooling in equimolar proportion

# Size selection

