---
title: "Using fastsimcoal"
author: "J. Ignacio Lucas Lledó"
date: "23/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The data

Here the goal is to use fastsimcoal2 to model the split of the ancestral fish population
that colonized the two lakes sampled in Norway. Thus, I need first to filter the VCF
to include only fish from the Arctic region. As usual, we start from the original, dirty
VCF, we remove unwanted individuals and then filter sites again. In order to determine
what individuals are from the Arctic lakes, I would use the original file. Note that in
folder `2022-05-04` we created a file, `2022-05-04/TheWorst.txt`, with the identifiers of
fish with the least data, that should be removed. They include some Arctic fish.

```{bash filteringFish}
VCF='../2022-04-22/assem2_outfiles/assem2.vcf'
REMOVABLE='../2022-05-04/TheWorst.txt'
FISHDATA='../../data/Fish_clean.tsv'

if [ ! -e Arctic.txt ]; then
   cut -f 1,4 $FISHDATA | grep "Arctic" | cut -f 1 > Arctic.txt
fi

if [ ! -e arctic.recode.vcf ]; then
   if [ ! -e dirty.recode.vcf ]; then
      vcftools --vcf $VCF \
               --out dirty \
               --keep Arctic.txt \
               --remove $REMOVABLE \
               --recode \
               --recode-INFO-all
   fi
   vcftools --vcf dirty.recode.vcf \
            --out arctic \
            --thin 500 \
            --remove-indels \
            --mac 1 \
            --max-alleles 2 \
            --max-meanDP 1000 \
            --max-missing-count 12 \
            --recode --recode-INFO-all
   if [ -e dirty.recode.vcf ]; then rm dirty.recode.vcf; fi
fi
```

I note that `vcftools --max-missing-count 12` is more conservative than expected. It does
not "exclude sites with more than this number of missing genotypes over all individuals",
but it seems to remove sites with more than this number of missing **alleles** over all
individuals. In this case of diploid genotypes, it removes individuals with more than 6
missing genotypes! I counted the number of reference, alternative and missing alleles in the
following way:

```{bash counting}
if [ ! -e arctic.recode.alleleCount.txt ]; then
   gawk 'BEGIN{
      print "#REF\tALT\tMissing\tTotal"
   }(/^[^#]/){
      delete F
      for (i=10;i<=NF;i++) {
         split($i, A, /:/)
         split(A[1], B, /\//)
         F[B[1]]++
         F[B[2]]++
      }
      print F["0"] "\t" F["1"] "\t" F["."] + 0 "\t" F["0"] + F["1"] + F["."]
   }' arctic.recode.vcf > arctic.recode.alleleCount.txt
fi
```

Note that when filtering the VCF, I use `--mac 1`, while we had used `--mac 2` in other
occasions. The reason is to avoid biases in the SFS. However, I am afraid the SFS will
necessarily be biased, due to ascertainment and downsampling. First, sites are selected for
being variable, and indeed all sites in `arctic.recode.vcf` have at least one allele different
from the rest. Thus, the zero category in the SFS should be empty. However, to deal with
missing data we will downsample the populations with `easySFS` (see below). That is, instead
of estimating the SFS for the 54 fish from Langfjordvatn and the 60 fish from Suopatjavri,
we estimate an SFS for a theoritcal sample of, say 50 and 50 fish, with no missing data. The
problem is that then, because the original VCF includes sites where only one individual
is heterozygous, when downsampling we expect those sites to be monomorphic. The zero-category
in the SFS becomes populated, but in a very biased way. Because I don't know how this bias
will affect the inferece, I need to estimate two different SFSs: with and without downsampling.

## PCA

Just to make sure that the lake is a good criteria to separate populations in the Arctic region.

```{r pca}
suppressMessages(library('ape'))
suppressMessages(library('pegas'))
suppressMessages(library('seqinr'))
suppressMessages(library('ggplot2'))
suppressMessages(library('adegenet'))
library('MetBrewer')
library(kableExtra)

arc <- read.vcf('arctic.recode.vcf', from = 1, to = 100000, quiet = TRUE)
arc <- as.data.frame(arc)
for (i in 1:dim(arc)[2]) arc[,i] <- as.character(arc[,i])
arc[arc == './.'] <- NA

arc.gi <- df2genind(arc, sep = '/', type = 'codom', ploidy = 2)
sum(is.na(arc.gi@tab))

X <- tab(arc.gi, NA.method = 'mean')
pca3 <- dudi.pca(X, scale=FALSE, scannf=FALSE, nf=3)
barplot(100 * pca3$eig[1:15] / sum(pca3$eig), main = 'PCA eigenvalues',
        col = met.brewer('Paquin', 15), ylab = "% of variance explained")
PercentVar <- head(100 * pca3$eig / sum(pca3$eig))

DATA <- '../../data/Fish_clean2.tsv'
Fish <- read.table(DATA, header = TRUE, sep = '\t')
row.names(Fish) <- Fish$Fish_code
arc.gi@pop <- droplevels(Fish[row.names(arc.gi@tab), 'Lake'])
col  <- met.brewer('Hiroshige', 2)  
#colt <- transp(col, 0.6)
par(mfrow = c(2,2))
plot(pca3$li$Axis1, pca3$li$Axis2,
     xlab = sprintf("PC 1, %.1f%%", PercentVar[1]),
     ylab = sprintf("PC 2, %.1f%%", PercentVar[2]),
     col = col[arc.gi@pop], pch = 16, cex = 1.5)
plot(pca3$li$Axis3, pca3$li$Axis2,
     xlim = c(max(pca3$li$Axis3), min(pca3$li$Axis3)),
     xlab = sprintf("PC 3, %.1f%%", PercentVar[3]),
     ylab = sprintf("PC 2, %.1f%%", PercentVar[2]),
     col = col[arc.gi@pop], pch = 16, cex = 1.5)
plot(pca3$li$Axis1, pca3$li$Axis3,
     xlab = sprintf("PC 1, %.1f%%", PercentVar[1]),
     ylab = sprintf("PC 3, %.1f%%", PercentVar[3]),
     col = col[arc.gi@pop], pch = 16, cex = 1.5)
plot(c(-40, 20), c(-20, 40), type = 'n', axes = FALSE, ann = FALSE)
legend(-40, y = 40, legend = levels(arc.gi@pop), fill = col, border= col)
par(mfrow = c(1,1))
```

The first component, accumulating more than 20% of the variation, separates fish
between lakes. Thus, I assume the two lakes are two populations or species.

## Estimation of the site frequency spectrum (SFS)

I have installed [easySFS](https://github.com/isaacovercast/easySFS), by Isaac Overcast.
To run it, I need to activate the conda environment called "easySFS". And I need a
population file. I assume the population file may include individuals other than those
in the VCF.

```{bash sfs,  engine.opts='-i'}
eval "$(command conda 'shell.bash' 'hook' 2> /dev/null)"
FISHDATA='../../data/Fish_clean2.tsv'
conda activate easySFS
if [ ! -e popmap.txt ]; then
   # Using lakes as population, because I only have Arctic fish in the VCF.
   if [ ! -e SamplesInVCF.txt ]; then
      grep "#CHROM" arctic.recode.vcf | \
      gawk '{for (i = 10; i <= NF; i++) print $i}' > SamplesInVCF.txt
   fi
   cut -f 1,2 $FISHDATA | tail -n +2 | grep -F -f SamplesInVCF.txt > popmap.txt
fi

if [ ! -d downsample ]; then
   if [ ! -e projection_preview.txt ]; then
      easySFS -a -i arctic.recode.vcf -p popmap.txt --preview > projection_preview.txt
   fi
   easySFS -a -i arctic.recode.vcf -p popmap.txt --proj 96,110 -o downsample
fi

if [ ! -d fullsample ]; then
   easySFS -a -i arctic.recode.vcf -p popmap.txt --proj 108,120 -o fullsample
fi
```

```{r plotSFS}
joint <- as.matrix(read.table('fullsample/fastsimcoal2/arctic_jointMAFpop1_0.obs',
                              skip = 1, header = TRUE))
Lang  <- t(as.matrix(read.table('fullsample/fastsimcoal2/Langfjordvatn_MAFpop0.obs',
                                skip = 1, header = TRUE)))
Suo   <- t(as.matrix(read.table('fullsample/fastsimcoal2/Suopatjavri_MAFpop0.obs',
                                skip = 1, header = TRUE)))
plot(Lang[1:55], type = 'l', xlab = 'Allele frequency',
     ylab = 'Number of SNPs', main = 'Langfjordvatn')
plot(Suo[1:60], type = 'l', xlab = 'Allele frequency',
     ylab = 'Number of SNPs', main = 'Suopatjavri')
library("RColorBrewer")
col <- colorRampPalette(brewer.pal(10, "RdYlBu"))(256)
heatmap(log(joint + 1), Rowv = NA, Colv = NA, xlab = 'Langfjordvatn',
        ylab = 'Supatjavri', col = col, scale = 'none')
```

The site frequency spectra of the two lakes suggest that Langfjordvatn has much
more genetic diversity than Suopatjavri. This is in agreement with the ML phylogeny
of the individuals, where fish from Suopatjavri form a clade inside the clade formed
by Langfjordvatn fish.

The SFS obtained are *projections*, even when the number of chromosomes *subsampled*
is equal to the total number of available chromosomes. For example, there are 54
fish from Langfjordvatn (108 chromosomes) and 60 fish from Suopatjavri (120 chromosomes)
and when all of them were requested to the easysSFS program, we end up with much fewer
SNPs than there were originally: 19541 SNPs in Langfjordvatn; 22024, in Suopatjavri;
and only 16830 in the joint dataset. Originally, there were 31411 SNPs in the VCF.
Therefore, it is not surprising that the row- or column totals in the joint SFS are
different from the numbers of SNPs in each populations.

The joint SFS suggests a common recent history of the two populations: few SNPs are
common in one population and rare in the other. And the most abundant class of SNPs
are those the minor allele of which is at low frequency in both populations.

## Running Fastsimcoal2

### Template file

Below I transcribe a template file describing a model with two populations that split
a while ago, and might have exchanged some individuals at some time afterwards. The
two populations are the two lakes. Instead of using a migration matrix to represent the
admixture event, I will assume it was a single event. To simplify, in the following
template, I assume that population 1 (Suopatjavri, I think) is the one founded later and
receiving migrants from population 0.

```{bash template, eval=FALSE}
//Number of population samples (demes)
2
//Population effective sizes (number of genes)
NPOP0
NPOP1
//Sample sizes
96
110
//Growth rates
0
0
//Number of migration matrices
0
//historical event: time, source, sink, migrants, new size, new growth rate, migr. matrix.
3 historical event
T1 1 0 MIG     1 0 0
T2 1 1   0 0.001 0 0
T2 1 0   1     1 0 0
//Number of independent loci [chromosomes]
1 0
//Per chromosome: Number of linkage blocks
1
//per Block: data type, num loci, rec. rate and mut rate + optional parameters
FREQ 1 0 1.0e-8 OUTEXP
```

### Estimation file

```{bash ranges, eval=FALSE}
// Priors and rules file
// *********************

[PARAMETERS]
//#isInt? #name #dist. #min #max
//all N are in number of haploid individuals
1 NPOP0 unif    100 1000000 output
1 NPOP1 unif    100 1000000 output
1 T1    logunif  10   10000 output
0 MIG   unif      0     0.5 output
1 T2    unif     T1   50000 output

[COMPLEX PARAMETERS]
```

I noticed the program fails without an error message (other than "core dumped")
if "[COMPLEX PARAMETERS]" is not present in the file. Also, I suspect that the
limits of log-uniformly-distributed variables are limits in the log scale.

### Execution to estimate the parameters of model 0

In order to estimate the parameters, the data (observed SFS) must be in the
same folder, and with a specific name. The name of the input file encodes the
kind of SFS being used: derived allele frequencies (DAF) or minor allele frequencies
("folded" SFS or MAF); one, two or more populations

```{bash estimate}
for run in $(seq 1 100); do
   rundir=$(printf "model0/run%03i" $run)
   if [ ! -d $rundir ]; then
      mkdir $rundir
   fi
   if [ ! -e $rundir/arctic.tpl ]; then
      cp model0/arctic.tpl $rundir/arctic.tpl
   fi
   if [ ! -e $rundir/arctic.est ]; then
      cp model0/arctic.est $rundir/arctic.est
   fi
   if [ ! -e $rundir/arctic_jointMAFpop1_0.obs ]; then
      ln -sr downsample/fastsimcoal2/arctic_jointMAFpop1_0.obs \
             $rundir/arctic_jointMAFpop1_0.obs
   fi
   if [ ! -e $rundir/arctic.par ]; then
      cd $rundir
      fsc27 -t arctic.tpl \
            -n 10000 \
            -m \
            -e arctic.est \
            -M \
            -L 40 \
            -c 0 \
            -q &> fsc27.log
      cd ../..
   fi
done

if [ ! -e model0/summary.txt ]; then
   echo -e "NPOP0\tNPOP1\tT1\tMIG\tT2\tMaxEstLhood\tMaxObsLhood" > model0/summary.txt
   find model0 -name arctic.bestlhoods -exec cat '{}' \; | \
   grep -v "^NPOP0" >> model0/summary.txt
fi
```

## Results

The absolute value of the population size estimates depend on the
predetermined mutation rate. And estimates of times of events depend
in turn on population sizes. If I express times relative to population
sizes, I may understand this better. For example, the time to an assumed
migration event from Langfjordvatn (population 0) to Suopatjavri (populatin 1)
can be expressed in units of coalescence in Suopatjavri by dividing T1
by NPOP1. Recall population sizes are expressed as chromosome numbers, as if
populations were haploid. Thus, a number of generations equal to the population
size is the expected time of coalescence of two randomly picked genes.
The time T2 when Suopatjavri's population started can also be expressed in
the same units.

```{r results}
library(ggplot2)
model0 <- read.table('model0/summary.txt', header = TRUE)
model0$T2c <- model0$T2 / model0$NPOP1
model0$T1c <- model0$T1 / model0$NPOP1
ggplot(data = model0, mapping = aes(x = T1c, y = T2c, color = MaxEstLhood)) +
   geom_point() + xlim(0, 0.015) + ylim(0.25, 0.75)
# Using xlim() and ylim() to focus in the region with highest likelihood.
```

Thus, it seems as if Suopatjavri's population started around half a coalescence
unit ago. That is, half the time expected for two random lineages to coalesce.
Quite recently. And the assumed migration event is even more recent.

Let's see how the two populations compare.

```{r relativeSize}
model0$N0overN1 <- model0$NPOP0 / model0$NPOP1
ggplot(data = model0, mapping = aes(x = N0overN1, y = MaxEstLhood)) +
   geom_point() + xlim(50,300) + ylim(-97000, -96000)
# I use xlim() and ylim() to focus on the most likely values.
```

So, it seems that Langfjordvatn maintains a population around 100 times
larger than that in Suopatjavri.

## Session Information
```{r sessionInfo}
sessionInfo()
```
