---
title: "insilico_digest"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using SimRAD to in silico digestion of the reference genome

To use SimRAD in R, two aditional packages from Bioconductor are necessary: "Biostrings" and "Shortread". Here, we perform a double-digest RADseq with a reference genome from Coregonus genus. We tested four restriction enzymes in pairs: PstI - MspI and SbfI - EcoRI.

### Install all necessary packages and load the reference genome 
```{r message=FALSE}

# install.packages("SimRAD")
# if (!requireNamespace("BiocManager", quietly = TRUE))
# install.packages("BiocManager")
# BiocManager::install("Biostrings")
# BiocManager::install("ShortRead")

library(SimRAD)

refseq <- ref.DNAseq("~/cophyhopa/data/reference.fa", subselect.contigs = TRUE, prop.contigs = 0.25)

```
### Define restriction enzymes 
The enzymes are defined according to the recognized restriction sites: 5 prime side and 3 prime side.

```{r}
# Define the restriction enzyme recognition pattern

# Restriction Enzyme SbfI
cs_5p1 <- "CCTGCA"
cs_3p1 <- "GG"
# Restriction Enzyme EcoRI
cs_5p2 <- "G"
cs_3p2 <- "AATTC"

# Restriction Enzyme 3 PstI
cs_5p3 <- "CTGCA"
cs_3p3 <- "G"
# Restriction Enzyme 4 MspI
cs_5p4 <- "C"
cs_3p4 <- "CGG"
```

### Test in silico digest with enymes 1 and 2 
Digestion with restriction enzymes 1 and 2 is simulated. Then those fragments flanked by two restriction enzyme sites are selected, one end with enzyme 1 and other end with enzyme 2 (type = "AB+BA"), during the library construction. Coming up next, those fragments within a specific range are selected. This step simulates the typical size selection step of ddRADseq. 

```{r warning=FALSE}
simseq.dig <- insilico.digest(refseq, cs_5p3, cs_3p3, cs_5p4, cs_3p4, verbose=TRUE)

# selection of AB type of fragments
simseq.sel <- adapt.select(simseq.dig, type="AB+BA", cs_5p3, cs_3p3, cs_5p4, cs_3p4)
# number of selected fragments: 
length(simseq.sel)

# wide size selection (500-1000bp):
wid.simseq <- size.select(simseq.sel, min.size = 500, max.size = 1000, graph=TRUE, verbose=TRUE)

# narrow size selection (600-900bp):
nar.simseq <- size.select(simseq.sel, min.size = 600, max.size = 900, graph=TRUE, verbose=TRUE)

# comparison of fragments size range (including too all fragments) 
boxplot(list(width(simseq.sel), width(wid.simseq), width(nar.simseq)), names=c("All fragments", "Wide size selection", "Narrow size selection"), ylab="Locus size (bp)")

# comparison of fragments size range (only with wide and narrow fragments) 
boxplot(list(width(wid.simseq), width(nar.simseq)), names=c("Wide size selection", "Narrow size selection"), ylab="Locus size (bp)")

```



