---
title: "Filtering and merging with VSEARCH"
author: "J. Ignacio Lucas Lledó"
date: "10/1/2022"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The program `vsearch` can filter, denoise, dectect chimeras and merge paired
end reads [@Rognes2016]. In what order should we process the reads? To remove
low quality reads is a good thing to do in the first steps, as well as removing
whole samples the sequencing of which may not have worked. That way, we don't
waste time processing low quality data.

Chimeric reads may be detected, and could then be removed as well. But it makes
sense to attempt merging before chimera detection, since it is the whole fragment
that might be chimeric, rather than only forward or only reverse reads, right?
Plus, chimera detection requires dereplicated, size-sorted fasta input.

Merging paired reads could be performed before filtering, because merging takes
quality into account and updates the quality of the overlapping bases. Plus,
VSEARCH does not seem to be able to filter paired-end reads without completely
messing up their order in the files. Data Carpentry recommends Trimmomatic for
that.

## Merging

Ideally I should combine fastq files from the two sequencing runs. This is
important for subsequent steps, like clustering. The strategy will be:

- Merge paired reads independently for the two sequencing runs.
- Combine outputs and distribute them in two folders: merged and paired.
- Compress the combined fastq files.

Because there seems to be plenty of data, I will ignore the sequences from
HOP files, even if some of them can be assigned to their original sample.

```{bash merging}
FASTQDIR1='../../data/fastq'
FASTQDIR2='../../data/fastr'
if [ ! -d logs ]; then mkdir logs; fi
if [ ! -d merged ]; then mkdir merged; fi
if [ ! -d paired ]; then mkdir paired; fi
if [ ! -d run1 ]; then mkdir run1; fi
if [ ! -d run2 ]; then mkdir run2; fi

for fwdRead in $(ls -1 $FASTQDIR1 | grep '_R1_' | grep -v HOP); do
   revRead=$(echo $fwdRead | sed 's/_R1_/_R2_/g')
   sample=$(echo $fwdRead | cut -d '_' -f 1)
   if [ ! -e merged/$sample.fq.gz ]; then
      if [ ! -e run1/$sample.merged.fastq ]; then
         vsearch --fastq_mergepairs $FASTQDIR1/$fwdRead \
                 --reverse $FASTQDIR1/$revRead \
                 --fastqout run1/$sample.merged.fastq \
                 --fastqout_notmerged_fwd run1/$sample.R1.fastq \
                 --fastqout_notmerged_rev run1/$sample.R2.fastq \
                 --fastq_allowmergestagger \
                 --fastq_maxdiffs 1 \
                 --log logs/$sample.1.merging.log \
                 --quiet --threads 20 &
      fi
      if [ ! -e run2/$sample.merged.fastq ]; then
         vsearch --fastq_mergepairs $FASTQDIR2/$fwdRead \
                 --reverse $FASTQDIR2/$revRead \
                 --fastqout run2/$sample.merged.fastq \
                 --fastqout_notmerged_fwd run2/$sample.R1.fastq \
                 --fastqout_notmerged_rev run2/$sample.R2.fastq \
                 --fastq_allowmergestagger \
                 --fastq_maxdiffs 1 \
                 --log logs/$sample.2.merging.log \
                 --quiet --threads 20 &
      fi
      wait
      cat run1/$sample.merged.fastq run2/$sample.merged.fastq | gzip > merged/$sample.fq.gz
      rm run1/$sample.merged.fastq run2/$sample.merged.fastq
      if [ -e run1/$sample.R1.fastq ]; then
         cat run1/$sample.R1.fastq run2/$sample.R1.fastq | gzip > paired/$sample.R1.fq.gz
         cat run1/$sample.R2.fastq run2/$sample.R2.fastq | gzip > paired/$sample.R2.fq.gz
         rm run1/$sample.R1.fastq
         rm run1/$sample.R2.fastq
         rm run2/$sample.R1.fastq
         rm run2/$sample.R2.fastq
      fi
   fi
done
rm -r run1
rm -r run2
```

I shoul have redirected standard output. Now I miss the statistics on how many pairs
got merged and all that. The merging process took two days, and I don't want to repeat it.
It is faster to now read all the files with `awk` to count and measure the length of the
merged or not-merged pairs.

```{bash mergeStats}
if [ ! -e merged.stats.txt ]; then
   if [ ! -d merged.stats ]; then mkdir merged.stats; fi
   NUMSAMPLES=$(ls -1 merged | wc -l)
   NUMTHREADS=40
   # Make sure we split samples evenly among threads:
   while [ $(( $NUMSAMPLES % $NUMTHREADS )) -ne 0 ]; do
      NUMTHREADS=$(( NUMTHREADS + 1))
   done
   NUMBATCHES=$(( NUMSAMPLES / NUMTHREADS))
   for i in $(seq 1 $NUMBATCHES); do
      LAST=$((  NUMTHREADS * i ))
      for sample in $(ls -1 merged | cut -d "." -f 1 | head -n $LAST | tail -n $NUMTHREADS); do
         echo $sample > merged.stats/$sample.stats.txt
         gunzip -c merged/$sample.fq.gz | \
         gawk '(NR % 4 == 2){
            F[(length($1))]++
         }END{
            for (f = 10; f <= 292; f++) print F[f] + 0
         }' >> merged.stats/$sample.stats.txt &
      done
      wait
   done
   for i in $(seq 1 $NUMBATCHES); do
      LAST=$((  NUMTHREADS * i ))
      for sample in $(ls -1 merged | cut -d "." -f 1 | head -n $LAST | tail -n $NUMTHREADS); do
         gunzip -c paired/$sample.R1.fq.gz | wc -l | \
         gawk '{print $1/4}' >> merged.stats/$sample.stats.txt &
      done
      wait
   done
   echo 'Length' > z1
   for i in $(seq 10 293); do echo $i >> z1; done
   paste z1 merged.stats/* > merged.stats.txt
   rm z1
   rm -r merged.stats
fi
```

The `merged.stats.txt` file contains a *histogram* of DNA fragment lengths for every sample.
The lengths range from 10 to 293. Because during merging I required the overlap to be at least
10 and original paired reads were 151 bases, $2\times 151-10 = 292$ is the maximum length of
merged read pairs. The *number of reads* assigned to the following length value (293) is the
number of non-merged read pairs, the length of which we ignore.

```{r lengthDist, fig.width=10, fig.height=10}
library(tidyr)
library(gridExtra)
library(ggplot2)
LengthHist <- read.table('merged.stats.txt', header = TRUE)
LongHist <- pivot_longer(LengthHist, 2:281, names_to = 'Sample', values_to = 'NumReads')
Filters <- lapply(unique(substr(LongHist$Sample, 1, 3)),
                  function(x) startsWith(LongHist$Sample, x))
names(Filters) <- unique(substr(LongHist$Sample, 1, 3))

plots <- lapply(names(Filters),
                function(x) {
                   ggplot(LongHist[Filters[[x]],],
                          aes(x=Length, y=NumReads, color=Sample)) +
                      geom_line() + scale_y_log10() + guides(color = 'none') +
                      ggtitle(x)})
grid.arrange(grobs = plots, nrow = 3)
```

Two remarkable features: the distributions of merged fragment lengths are extremely
similar, with all peaks matching among samples. This is very good because it
means that the short DNA fragments that sneaked into the analysis are the same ones
in all samples. And second, the last peak is the reason to use a log scale and 
indicates that in all samples there are still many reads from fragments longer than
292 base pairs, which is good. Let's see the proportions of merged and non-merged
pairs.

```{r proportions}
library(RColorBrewer)
Merging <- data.frame(
   Sample = names(LengthHist)[2:281],
   Merged = colSums(LengthHist[1:(dim(LengthHist)[1] - 1), 2:281]),
   Paired = as.numeric(LengthHist[dim(LengthHist)[1], 2:281]),
   Total  = colSums(LengthHist[, 2:281])
)
barplot(t(as.matrix(Merging[order(Merging$Total, decreasing = TRUE), 2:3])),
        xaxt = 'n', border = NA, space = 0,
        col = brewer.pal(3, 'Set2')[c(1,2)])
legend(180, 2e+07, c('Merged', 'Not merged'), fill = brewer.pal(3, 'Set2')[c(1,2)])

Merging$PropMerged <- Merging$Merged / Merging$Total
ggplot(Merging, aes(x = Total, y = PropMerged)) + geom_point() + geom_smooth(method = 'lm')
```

About 50% of read pairs got merged. The number of non-merged reads is high
enough to justify working separately with both kinds of reads.

The purpose of clustering now is to count the number of different loci, and
we do not need to include reverse reads in that analysis, because the same
clustering is expected using either forward or reverse reads. And artificially
merging them does not seem a good idea.

## Filtering

## Clustering

## Session Info

```{r sessionInfo}
sessionInfo()
```

# References
